# Simple Linear Regression {#slr}

This chapter contains content from Sleuth Chapters 7 and 8. 

## The variables
Suppose we have a quantitative response variable $y$ that we want to relate to an explantory (aka predictor) variable $x$. For now, we will assume that $x$ is also quantitative. 


## The model form {#slr-model}
This section describes the SLR model for a particular **population** of interest. Another way to frame the model is that it describes a hypothetical **data generating process (DGP)** that was used to generate the sample of data that we have on hand. 

Let $Y_i$ be the response from unit $i$ that has explanatory (aka predictor) value $x_i$. There are two **equivalent** ways to express the SLR model for $Y$:

- Conditional normal model: Given a value $x_i$, the response $Y_i$ follows a normal model with mean and SD given below:
$$
Y_i \mid x_i \sim N(\mu_{y\mid x} = \beta_0 + \beta_1 x_i, \sigma)
$$

- Mean + error: Statisticians are more likely to use a model specification that expresses $Y$ as a function of the **expected value/mean** of $Y$ plus an **error** term that models variation in responses around the mean:
\begin{equation}
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i \ \ \ \ \ \epsilon_i \sim N(0, \sigma)
\end{equation}

**Both** expressions of the SLR model above say the same thing: 

- **Linear Mean:** $\mu_{y\mid x} = E(Y \mid x) =  \beta_0 + \beta_1 x$ describes the population mean value of $Y$ given a predictor value $x$. This mean value varies linearly with $x$ and the population parameters are $\beta_0$ and $\beta_1$. 
- **Constant SD:** $SD(Y\mid x)=\sigma$ describes the SD of $Y$'s in the population around a given mean value $\mu_{y\mid x}$. The fact that this SD **does not** depend on the value of $x$ is called the contant variance, or homoscedastic, assumption. 
- **Normality:** The shape of population response values around $\mu_{y\mid x}$ is described by a normal distribution model.  

Finally, one last assumption is made for the SLR model:

- **Indepedence:** Given a predictor value of $x$, all responses $Y$ occur independently of each other. 

There are a total of **three parameters** in the SLR model: 

- the two mean parameters $\beta_0$ and $\beta_1$
- the SD parameter $\sigma$

### Interpretation

For a SLR model:

- $\beta_0$ is the mean response when the predictor value is 0 since $\mu_{y \mid 0} = \beta_0 + \beta_1(0) = \beta_0$.
- $\beta_1$ tells us how the mean response changes for a one unit increase in $x$

### Example: Woodpecker nests {#slr-model-ex}

We want to model nest depth (cm) in a tree cavity as a function of ambient air temperature (Celsius). Our SLR model for this relationships says that, given an ambient air temp $x_i$, a randomly selected nest will have a depth $Y_i$ that is modeled as
$$
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i \ \ \ \ \ \epsilon_i \sim N(0, \sigma)
$$
This means that depth is normally distributed with a mean of $\mu_{depth \mid temp} = \beta_0 + \beta_1 (temp)$ and a SD of $\sigma$. 

-----------

## Theory: Estimation {#slr-est}

Let's consider taking a random sample of size $n$ of responses and predictor values from our population (or DGP) for which the SLR model holds: $(x_1, Y_1), \dotsc, (x_n, Y_n)$. The notation here, $(x_i, Y_i)$ implies that the predictor value $x_i$ is fixed, but $Y_i$ is a **random variable** that is generated from the SLR model described in Section \@ref(slr-model). The **estimation** problem is that we need to use the sample of size $n$ to estimate the SLR model parameters $\beta_0$, $\beta_1$ and $\sigma$.

Once we observe a sample of size $n$, then we can use the SLR model to determine the probability of observing the sample. This probability, which depends on the actual model parameters $\beta_0$, $\beta_1$ and $\sigma$, is called a **likelihood function**. We plug the observed data into this function, then find the parameters values that **maximize** the function using calculus. For a SLR model, this process yields the following **maximum likelihood estimators (MLE)** of our parameters: 
$$
\hat{\beta}_1 = \dfrac{\sum_{i=1}^n (x_i - \bar{x})(Y_i - \bar{Y})}{\sum_{i=1}^n (x_i - \bar{x})^2} \ \ \ \ \ \hat{\beta}_0 = \bar{Y} - \bar{\beta}_1 \bar{x} \ \ \ \ \ \hat{\sigma} = \sqrt{\dfrac{\sum_{i=1}^n (Y_i - \hat{y}_i)^2}{n-2}}
$$
where $\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$ is the **predicted value** of $y$ given the value $x_i$.

### Sampling Distributions for SLR estimates

The sampling distribution of a model estimate ($\hat{\beta}_0$ or $\hat{\beta}_1$) is constructed by:

1.  fix a set of predictor values: $x_1, \dotsc, x_n$
2.  for each fixed $x_i$, generate a response $Y_i$ from $N(\beta_0 + \beta_1 x_i, \sigma)$
3. compute the MLE's $\hat{\beta}_0$ and $\hat{\beta}_1$ from the observed sample from (2)
4. repeat 2-3 lots of times, then the distribution of the estimates from part (3) show the sampling distribution of the slope or intercept estimate.

Using probability theory, we can show that the sampling distributions of both $\hat{\beta}_0$ and $\hat{\beta}_1$ are approximately normal when $n$ is "large enough" (thanks to the CLT):
$$
\hat{\beta}_i \sim N(\beta_i, SD(\hat{\beta}_i))
$$

- **Unbiased:** We see that the expected value (mean) of $\hat{\beta}_i$ is the parameter $\beta_i$, meaning it is an unbiased estimator. (It doesn't systematically over- or under-estimate the parameter of interest.)
- **Standard error:** We end up estimating the SD in the sampling distribution given above. The SEs for each mean parameter estimate are
$$
SE(\hat{\beta}_1) = \hat{\sigma}\sqrt{\dfrac{1}{(n-1)s^2_x}} \ \ \ \ \ SE(\hat{\beta}_0) = \hat{\sigma}\sqrt{\dfrac{1}{n} + \dfrac{\bar{x}^2}{(n-1)s^2_x}}
$$

-----------

## SLR model simulation {#slr-sim}

Download the Markdown of this activity: [.Rmd](https://kstclair.github.io/Math245/Rhandouts/Day3RegLineSim.Rmd).


### Simulation function
This chunk contains code that defines our function `reg.sim` that **simulates $n$ responses from a given regression model and given set of $n$ predictor values $x$**. I've excluded it from our compiled document so see the .Rmd file to take a look at how this was created. 

```{r, include=FALSE}
reg.sim <- function(x, beta0, beta1, sigma, grph=T)
{
  # function takes a sample of responses y generated from the given model/x values
  # returns estimated intercept and slope
  # x = explantory variable
  # beta0, beta1 = true model intercept and slope
  # sigma = true model SD
  # grph = print simulation graphs (y vs. x)? (default is true)
  n<- length(x)
  muy.x <- beta0 + beta1 * x

  # simulate some y's, then fit lm using these y's:
  y <- rnorm(n, muy.x,sigma)
  mylm <- lm(y~x)

  # plot data, LS line, true line
  if (grph){
    plot(x,y,xlab="x",ylab="simulated y values",pch=16, ylim=c(0,30))
    legend("topright",legend=c("Population line", "Sample line"),col=c("red","black"),lty=c(1,1),lwd=2)
    abline(mylm,lwd=2)
    abline(beta0,beta1,lwd=2,col="red")
    }
  return(list(b0=mylm$coefficients[[1]], b1=mylm$coefficients[[2]] ))
}
```

### Run the function once
Let's use the function from (1) above. We will use the $n=12$ temps (x-values) from the woodpeckers data and assume that the **true** model is:
$$\mu(y \mid x) = 20 - 0.4x$$ 
(red line below) with  $\beta_0=20$, $\beta_1=-0.4$, and $\sigma=2$. In the code below I use the `set.seed()` command to "fix" the random number generator so I get the same answer each time this is run (so my answer in the handout is reproduced each time this file is compiled).

```{r, fig.width=5,fig.height=3}
wpdata<- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
set.seed(77)  
reg.sim(x=wpdata$temp, beta0=20, beta1=-.4, sigma=2)
```

For this simulated sample (with seed of 77), the estimated regression line is $\hat{\mu}(y \mid x) = 20.889 - 0.468x$ (black line).

### Simulated sampling distribution for $\hat{\beta}_1$
We will now use the `replicate` command to generate 1000 different samples which create 1000 different estimates of $\beta_1$. A histogram of these estimates simulates the sampling distribution of estimated slope.  

**What is the shape of the sampling distribution? Where is the distribution centered? How variable are these estimated slopes.**

```{r, fig.width=5,fig.height=3.5 }
set.seed(7)  # just makes simulation reproducible
slopes<- replicate(1000,reg.sim(x=wpdata$temp, beta0=20, beta1=-.4, sigma=2,grph=F)$b1)
hist(slopes); abline(v=-0.4,col="red", lwd=2)
mean(slopes); sd(slopes)
```


### Are slope and intercept estimates correlated?
In regression, it is not unusual to be interested in estimating a **linear combination** of our model parameters. An easy example of such a combination of parameters is the mean response for a given value $x_0$ of the predictor: 
$$
\mu(y \mid x=x_0) = \beta_0 + \beta_1x_0
$$
For a specific example, we may want to estimate the mean response (depth) for a temp of $x_0=5$ degrees: $\mu(y \mid x=5) = \beta_0 + \beta_1 5$. If we don't know $\beta_0$ and $\beta_1$, then this mean parameter is a linear combination of two unknown parameters which we need to estimate. 

The natural estimate of this is just the estimated mean response: $\hat{\mu}(y \mid x=x_0)=\hat{\beta}_0 + \hat{\beta}_1x_0$. To assess how variable this estimate is (i.e. to get its SE) we need to understand how (if) the individual estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ are correlated. We can use a simulation to look at this issue by generating 1000 samples from our model and plotting each $(\hat{\beta}_0,\hat{\beta}_1)$ pair for each sample. 

**How are estimated slope and intercept associated? Any ideas why?**

```{r}
set.seed(7)  # this seed MUST match the seed used to get slopes!
intercepts<- replicate(1000,reg.sim(x=wpdata$temp, beta0=20, beta1=-.4, sigma=2,grph=F)$b0)
plot(intercepts,slopes); abline(h=-.4,col="red"); abline(v=20,col="red")
title("Estimated slopes and intercepts")
cor(intercepts, slopes) # correlation between estimates
cov(intercepts, slopes) # covariance between estimates
```




--------------------------------

## Inference for mean parameters {#slr-inf}

In intro stats, you used $t$ inference procedures for inference about population means since: (1) the sampling distribution of the sample mean was normally distributed and (2) we had to estimate its variability with a SE. The same goes for inference about the mean response parameters in a SLR model:
$$
t = \dfrac{\hat{\beta}_i - \beta_i}{SE(\hat{\beta}_i)} \sim t_{df=n-2}
$$
Use a t-distribution with $n-2$ degrees of freedom for inference about the mean parameters. The degrees of freedom are calculated as the sample size $n$ minus the number of terms in $\mu_{y \mid x}$ that you have to estimate with the data. 


### Confidence Intervals
To estimate either mean parameter with $C$% confidence, we have the general form
$$
\hat{\beta}_i \pm t^* SE(\hat{\beta}_i)
$$
where $t^*$ is the $(100-C)/2$ percentile from the t-distribution with $df=n-2$ degrees of freedom. 

### Hypothesis tests
We can test the hypothesis
$$
H_0: \beta_i = \beta^*_i
$$
with the following t-test statistic:
$$
t =\dfrac{\hat{\beta}_i - \beta^*_i}{SE(\hat{\beta}_i)}
$$
where $\beta^*$ is our hypothesized value of $\beta$ (intercept or slope). The t-distribution with $n-2$ degrees of freedom is used to compute the p-value that is appropriate for whatever $H_A$ is specified. 

The usual test results given by standard regression output tests whether a parameter value (intercept or slope) is equal to 0 vs. not equal to 0:
$$
H_0: \beta_i = 0 \ \ \ \ \ H_A: \beta_i \neq 0
$$
with a test stat of 
$$
t =\dfrac{\hat{\beta}_i - 0}{SE(\hat{\beta}_i)}
$$

----------------------------

## Inference for average or predicted response {#slr-inf2}

### Confidence intervals for $\mu_{y \mid x}$

Here we are interested in estimating not just a $\beta$ with confidence, but we want to estimate the **mean** response for a given value of $x_0$. For example, suppose we want to estimate the **mean** nest depth when the temp is $x_0=8$ degrees. This mean parameter of interest is then:
$$
\mu_{depth \mid temp=8} = \beta_0 + \beta_1 (8)
$$


Our parameter of interest is $\mu_{y \mid x_0} = \beta_0 + \beta_1 x_0$ where $x_0$ is known and $\beta$'s need to be estimated. The natural estimator is just the fitted equation:
$$
\hat{\mu}_{y \mid x_0} = \hat{\beta}_0 + \hat{\beta}_1 x_0
$$
As will *any* estimator, we can measure the variability of this estimator with a SE:
$$
SE(\hat{\mu}_{y \mid x_0}) = \hat{\sigma} \sqrt{\dfrac{1}{n} + \dfrac{(x_0 - \bar{x})^2}{(n-1)s^2_x}}
$$
**Note!** This SE **depends on $\pmb{x_0}$**! It is miminized when $x_0$ equals the mean predictor value $\bar{x}$ and it grows as $x_0$ gets further from $\bar{x}$. Estimation is most precise in the "middle" of the predictor range and becomes less precise at the "edges" (where we usually have less data). 

A 95% confidence interval for the mean response $\mu_{y \mid x_0}$ looks like
$$
\hat{\mu}_{y \mid x_0} \pm t^*_{df=n-2}SE(\hat{\mu}_{y \mid x_0}) 
$$

### Prediction intervals for new cases
To predict *one individual's future response* $Y$ for the predictor value $x_0$, we just use the  fitted equation:
$$
pred_{y \mid x_0} = \hat{\beta}_0 + \hat{\beta}_1 x_0
$$
(We use $pred_{y \mid x_0}$ to remind us which predictor value as used for prediction.) 

Recall that the SLR model assumes the $Y$ values when $x=x_0$ are normally distributed with mean $\mu_{y \mid x_0}$ and SD $\sigma$. The SE of this prediction at $x_0$ takes into account (1) uncertainty in estimating the *mean*  $\mu_{y \mid x_0}$  and (2) variation in $Y$'s around the mean response ($\sigma$):
$$
SE(pred_{y \mid x_0}) = \hat{\sigma} \sqrt{1 + \dfrac{1}{n} + \dfrac{(x_0 - \bar{x})^2}{(n-1)s^2_x}} = \sqrt{\hat{\sigma}^2 + SE(\hat{\mu}_{y \mid x_0})^2  }
$$
A 95% **prediction** interval for a future individual response at $x=x_0$ looks like
$$
pred_{y \mid x_0} \pm t^*_{df=n-2}SE(pred_{y \mid x_0})
$$

Predictions intervals feel and look similar to the mean response intervals above, but there is a very important conceptual difference: *prediction* means we are trying to "guess" at *one individual response* as opposed to the mean response of a large group of individuals. For example, if we want to understand nest depths for all nest build when temp is 8 degrees, then we care about estimating a fixed (but unknown) mean depth $\mu_{depth \mid temp=8}$. If we see a bird starting to build a nest at 8 degrees, then we care about predicting this one, randomly determined depth $Y$ using $pred_{depth \mid temp=8}$ and would use a prediction interval. 


--------------------------------

## Example: SLR model (day 3) {#slr-example1}
We will revisit the woodpecker nesting data first described in Section \@ref(slr-model-ex). 

### Load data    
Let's suppose these 16 nests are a random sample of all nests in the collection region. The command `head(dataname)` produces a view of the first 5 rows of data.

```{r, eval = TRUE}
wpdata <- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
head(wpdata)
dim(wpdata)
```
```{r, include = FALSE}
wpdata <- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
```

### EDA
Start with univariate exploration:

```{r, eval = TRUE}
summary(wpdata)
par(mfrow=c(1,2))
hist(wpdata$temp)
hist(wpdata$depth)
par(mfrow=c(1,1))  
```

Graphically explore the ("bivariate") relationship between temp and depth with a scatterplot using the command `plot(y,x)`. Here is the `plot` version (with `pch` point character changed to give filled circles):

```{r, fig.width=6, fig.height=4, eval = TRUE}
plot(depth ~ temp, data=wpdata, pch=16, xlab="air temperature (C)", 
     ylab="nest depth (cm)", main="woodpeckers scatterplot")
```

Here is the `ggplot2` version with `labs` added to change labels and title:
```{r, fig.width=6, fig.height=4, eval = TRUE}
library(ggplot2)
ggplot(wpdata, aes(x=temp, y = depth)) + geom_point() +
  labs(x = "air temperature (C)", y= "nest depth (cm)", title= "woodpeckers scatterplot")
```

### The least squares line (the estimated SLR model):
You fit the linear model with mean $\mu_{Y \mid x} = \beta_0 + \beta_1 x$ with the linear model function `lm(y ~ x, data=)`.

```{r, eval = TRUE}
wood.lm <- lm(depth~temp, data=wpdata)
wood.lm
```
```{r, include = FALSE}
wood.lm <- lm(depth~temp, data=wpdata)
wood.lm
```
We have mean parameter estimates:
$$
\hat{\beta}_0 = `r round(wood.lm$coefficients[1],3)` \ \ \ \ \hat{\beta}_0 = `r round(wood.lm$coefficients[2],3)`
$$

The object `wood.lm` is called a *linear model object* in R. We can add the regression line from this object to an existing *base R* plot of the data using the `abline` command.
```{r, fig.width=6, fig.height=4, eval = TRUE}
plot(depth ~ temp, data=wpdata, pch=16, xlab="air temperature (C)", 
     ylab="nest depth (cm)", main="regression of depth on temp")
abline(wood.lm)
```

The `ggplot2` package contains a `geom_smooth` geometry to add this SLR line:
```{r, fig.width=6, fig.height=4, eval = TRUE}
ggplot(wpdata, aes(x=temp, y = depth)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "air temperature (C)", y= "nest depth (cm)", title= "woodpeckers scatterplot")
```

With `geom_smooth`, you need to specify the type of method used to relate $x$ to $y$. Adding `se=FALSE` removes a confidence interval for the mean trend. Notice that the `aes` aesthetic given in the `ggplot` function specifies the x and y variables to plot. The `geom`'s that follow this part of the command use this `aes` to create points and an estimated line.   



### Inference for coefficients
The `summary` command is used for statistical inference for the intercept $\beta_0$ and slope $\beta_1$:

```{r, eval = TRUE}
summary(wood.lm)
```
This output gives the  mean parameter estimates in the `Estimate` column of the main table:
$$
\hat{\beta}_0 = `r round(wood.lm$coefficients[1],3)` \ \ \ \ \hat{\beta}_0 = `r round(wood.lm$coefficients[2],3)`
$$
The estimated model SD $\sigma$ is given by `Residual standard error`:
$$
\hat{\sigma} = `r round(summary(wood.lm)$sigma,3)`
$$


You should know how to verify (or find) the test stats and p-values for $\beta_0$ and $\beta_1$ given by the `summary` command if you are given the estimates and SEs. For the slope test $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$ we get a  $t$ test stat of

```{r, eval = TRUE}
(-0.34218 - 0)/0.05961
```

The p-value for this two sided test is the probability of being above +5.741 and below -5.741, or double the probability below -5.741 (since the t-distribution is symmetric around 0):

```{r, eval = TRUE}
pt(-5.741,10)
2*pt(-5.741,10)
```

A 95% confidence interval for the slope $\beta_1$ is computed from the t distribution with 10 degrees of freedom:

```{r, eval = TRUE}
qt(.975,10)
-.34218 + c(-1,1)*qt(.975,10)*.05961
confint(wood.lm)
```

The `confint` function gives the most accurate interval (no rounding error) but you need to know how to compute these CIs ``by hand."


### Additional `lm` information
The function `lm` creates a *linear model* object in R that has lots of information associated with it. Information includes the coefficient values, fitted values ($\hat{y}_i$), and residuals ($y_i - \hat{y}_i$):

```{r, eval = TRUE}
attributes(wood.lm)
wood.lm$coefficients
wood.lm$fitted.values   # predicted y values (y-hat)
wood.lm$residuals       # residuals for each data point
```

There are also functions that act on `lm` objects like
```{r, eval = TRUE}
fitted(wood.lm)
resid(wood.lm)
vcov(wood.lm)  # variance and covariance matrix of beta estimates
```


### `broom` package: Tidy `lm` output {#slr-broom}
The `broom` package contains functions that convert model objects (like a `lm` object) into "tidy" data frames. The `tidy` command summarizes model results:
```{r, eval = TRUE}
library(broom)
tidy(wood.lm)
```
Notice that the output object is called a "tibble" which is a type of data frame in R. 

Here we can add confidence intervals for the model parameters to the output:
```{r, eval = TRUE}
tidy(wood.lm, conf.int=TRUE)
```

The `augment` command augments the data frame used to create a `lm` with predicted values and residuals from the `lm` model:
```{r, eval = TRUE}
wpdata.aug <- augment(wood.lm)
head(wpdata.aug)
```
The `.fitted` column gives the fitted values $\hat{y}_i$ for each case and the `.resid` gives the residuals $y_i - \hat{y}_i$.


### Inference for the mean and predicted response
We can get a 95% confidence interval for the mean nest depth at 8 degrees Celsius $\mu(depth\mid temp=8)$  with the `predict` command with interval type `confidence` specified:

```{r, eval = TRUE}
predict(wood.lm, newdata = data.frame(temp=8), interval = "confidence", se.fit=T)
```
The argument `se.fit=T` provides the SE of the mean response estimate. The estimated mean depth of all nests built at 8 degrees is 17.38 cm with a SE of 0.697 cm. We are 95% confident that the mean depth at 8 degrees is between 15.83 to 18.9 cm. 

You can verify the computation of the SE and CI using summary stats for `temp` and the estimated parameter values from the `summary` command:
$$
\hat{\mu}_{temp \mid 8} = \hat{\beta}_0 + \hat{\beta}_1 (8) = `r round(wood.lm$coefficients[1],3)` + (`r round(wood.lm$coefficients[2],3)` )(8) = `r round(wood.lm$coefficients[1] + wood.lm$coefficients[2]*8,3)`
$$
$$
SE(\hat{\mu}_{depth \mid 8}) = \hat{\sigma} \sqrt{\dfrac{1}{n} + \dfrac{(x_0 - \bar{x})^2}{(n-1)s^2_x}} = `r round(summary(wood.lm)$sigma, 3)` \sqrt{\dfrac{1}{12} + \dfrac{(8-`r round(mean(wpdata$temp),3)`)^2}{(12-1)`r round(sd(wpdata$temp),3)`^2}} = 0.6972
$$

```{r, eval = TRUE}
nrow(wpdata)
mean(wpdata$temp)
sd(wpdata$temp)
mn.est.se <- 2.33453*sqrt(1/12 + (8-11)^2/((12-1)*11.80909^2))
mn.est.se
mn.est <- 20.12228 -0.34218*8
mn.est
mn.est + c(-1,1)*qt(.975,10)*mn.est.se
```

You can also include more than one predictor value temp in this function:
```{r, eval = TRUE}
predict(wood.lm, newdata = data.frame(temp=c(8,20)), interval = "confidence", se.fit=T)
```

We can get a 95% prediction interval for the depth of one depth constructed at 8 degrees Celsius $pred_{depth \mid temp=8}$  with the `predict` command with interval type `prediction` specified:

```{r, eval = TRUE}
predict(wood.lm, newdata = data.frame(temp=8), interval = "prediction")
```
We are 95% confident that a new nest built at 8 degrees will have a depth between 11.96 to 22.81 cm. 


R does not give us the SE for prediction $SE(pred(Y \mid x_0)) = \sqrt{\hat{\sigma}^2 + SE(\hat{\mu})^2}$ so we need to compute it by hand if we want its value:
$$
SE(pred_{depth \mid 8}) = \sqrt{\hat{\sigma}^2 + SE(\hat{\mu})^2} =  \sqrt{`r round(summary(wood.lm)$sigma, 3)`^2 + 0.6972^2} = 2.436
$$

```{r, eval = TRUE}
se.pred <- sqrt(2.33453^2 + 0.6972406^2)
se.pred
```

The predicted depth of one nest build at 8 degress is 17.38 cm with a SE of 2.44 cm. The 95% prediction interval produced above can be verified as follows:

```{r, eval = TRUE}
yhat <-  20.12228 -0.34218*8
yhat
yhat + c(-1,1)*qt(.975,10)*se.pred
```


### Adding confidence bands to a scatterplot
The `geom_smooth` function in `ggplot2` adds a 95% confidence interval for $\mu_{y \mid x}$ around the estimated mean line:
```{r, eval = TRUE}
ggplot(wpdata, aes(x=temp, y = depth)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(x = "air temperature (C)", y= "nest depth (cm)", title= "woodpeckers scatterplot with mean confidence interval")
```

Adding prediction interval bands takes slightly more work. First, create a new version of the data set that includes prediction intervals for each case in the data:
```{r, eval = TRUE}
wpdata.pred <- data.frame(wpdata, predict(wood.lm, interval="prediction"))
head(wpdata.pred)
```

Then add a `geom_ribbon` layer to the previous plot, with `ymin` and `ymax` determined by the prediction interval's lower (`lwr`) and upper (`upr`) bounds. The `fill` arguments in both layers below are not really needed, but they are used here to provide a legend label for the plot:
```{r, eval = TRUE}
ggplot(wpdata.pred, aes(x=temp, y = depth)) + 
  geom_point() +
    geom_ribbon(aes(x=temp, ymin = lwr, ymax = upr, fill = "prediction"), alpha = .1) +
  geom_smooth(method = "lm", aes(fill = "confidence"), alpha = .4) +
  labs(x = "air temperature (C)", y= "nest depth (cm)", 
       title= "woodpeckers scatterplot", fill = "Type")
```


### Tools for displaying your model
As described in the Formatting Tables in Markdown (Section \@ref(markdown-tables)), you can use the package `stargazer` to create a nice table of model results in your pdf. The entire R chunk to do this in a `pdf` doc format is shown below (but not evaluated in this html book). You will need to add the R chunk option `results='asis'` to get the table formatted correctly. I also include the `message=FALSE` option in the chunk below that runs the library command to suppress the automatic message created when running the `library` command with `stargazer`. 


````markdown
`r ''````{r, results='asis', message=FALSE}
library(stargazer)
stargazer(wood.lm, header=FALSE, single.row = TRUE, title = "SLR of depth on temp")
```
````
````

The `kable` function (from the `knitr` package) works well in all output environments (e.g. pdf, html, word). The input to this function needs to be a data frame, so we can use the `tidy` version of our model summary:

```{r, eval = TRUE}
library(knitr)
kable(tidy(wood.lm, conf.int = TRUE), digits = 3)
```



--------------------------------

## Checking model assumptions and fit {#slr-assump}

Recall the SLR model assumptions from \@ref(slr-model):

$$
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i \ \ \ \ \ \epsilon_i \sim N(0, \sigma)
$$

- **Linearity:** The mean response varies linearly with $x$.
- **Constant SD:** $SD(Y\mid x)=\sigma$ describes the SD of $Y$'s in the population around a given mean value $\mu_{y\mid x}$. An **equivalent** statement of this assumption is that the model *errors* should not be associated with $x$. 
- **Normality:** The shape of population response values around $\mu_{y\mid x}$ is described by a normal distribution model.  
- **Indepedence:** Given a predictor value of $x$, all responses $Y$ occur independently of each other. An **equivalent** statement of this assumption is that the model *errors* are independent. 

### Residuals 
**If** all four model assumptions are met, then our model errors $\epsilon_i$'s will be independent and distributed like $N(0,\sigma)$. Now we can't actually "see" the model errors unless we know the true parameter values in the population since
$$
\epsilon_i = y_i - (\beta_0 + \beta_1 x_i)
$$

The closest thing we have to the model errors are the fitted model residuals computed using the **estimated** model parameters:
$$
r_i = y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_i)
$$
Residuals for a fitted model are a **diagnostic** tool to help check whether a model fit to data seems to match the true model form that generated the data.

### Residual plot: linearity and constant variance

A **residual** plot is constructed by plotting $r_i$ (y-axis) against $x_i$ (x-axis). A horizontal reference line at $y=0$ is usually added (since the mean residual value is always 0). 

- **Linearity**: This assumption is met if, at each x-value,  you see similar scatter of points (residuals) above and below the 0-reference line.
- **Constant variance**: This assumption is met if you see a similar magnitude of point scatter around the 0-reference line as you move along the x-axis. 

A residual plot that meets both these conditions is called a **null** plot. 

Watch out for

- **curvature** which suggests the mean function relating $y$ and $x$ may not be linear
- **nonconstant variation** which is seem in "fan" shaped plots
- **outliers** which can have a large influence on the fitted model

#### Example: Residual plot

Let's revisit the woodpecker nesting depth model and use the `broom` package to add residuals to the `wpdata` data frame (Section \@ref(slr-broom)):

```{r}
# model fit above in Section 2.7 example
wpdata<- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
wood.lm<- lm(depth~temp, data=wpdata)
library(broom)
wpdata.aug <- augment(wood.lm)
head(wpdata.aug)
```

The residual plot will put the predictor  `temp` on the x-axis and `.resid` on the y-axis:
```{r}
library(ggplot2)
ggplot(wpdata.aug, aes(x = temp, y = .resid)) +
  geom_point()  +
  geom_hline(yintercept = 0, linetype= "dashed")
```

Don't forget to use the augmented data frame `wpdata.aug` that contains the residuals. The layer `geom_hline` adds the horizontal reference line at 0. 

**Interpretation:** There are no majors trends seen in this residual plot. Generally, it is hard to prove or disprove model assumptions when we only observe 12 data points!


### Residual normal QQ plot

A **normal QQ plot** for a variable plots observed **q**uartiles against the theoretical **q**uartiles from a normal model. If these points follow a line then the data is approximately normal. Here is a general guide for interpreting non-linear trends that indicate a non-normal distribution:

- Concave up: the distribution is right skewed
- Concave down: the distribution is left skewed
- S-shaped: the distribution is symmetric but the tails are either too short (not enough variation) or too long (too much variation) to be normally distributed.
- More help! Check out [this website](https://seankross.com/2016/02/29/A-Q-Q-Plot-Dissection-Kit.html) for a deeper discussion of the interpretation of normal QQ plots

We can check the **normality assumption** by plotting residuals with a normal QQ plot. You can also use a histogram of residuals to help interpret the normal QQ plot. 

#### Example: Residual normal QQ plot

Back to the augmented woodpecker data set. A histogram of the residuals shows a slightly right skewed distribution.
```{r}
hist(wpdata.aug$.resid)
```

A quick way to get a normal QQ plot of residuals is to plug the `lm` object into the `plot` command and request plot number 2:
```{r}
plot(wood.lm, which = 2)
```


Alternatively, you could use a `ggplot`. The aesthetic used for a QQ plot in a `ggplot` is `sample = variable`, then `geom_qq` and `geom_qq_line` are the layers used:
```{r}
ggplot(wpdata.aug, aes(sample = .resid)) +
  geom_qq() + 
  geom_qq_line()
```

**Interpretation:** The QQ plot also suggests a slightly longer right tail because it is (sort of) concave up. But, the QQ plot also clearly shows that this feature could just be due to two cases with residual values that are slightly higher than all others. Again, because there are only 12 data points these two cases could just be due to chance and we can conclude that there aren't any strong indications of a non-normal distribution. 

### Independence

Independence of errors is probably the hardest assumption to check because it often depends on how the data was collected. Two common scenarios that lead to data that violates this assumption are

- **temporal correlation:** This means correlation of errors because measurements were collected over time. Responses (and errors) measured close in time are more likely to be similar in value that responses measured further apart in time. For example, daily temperature readings in Northfield are temporally correlated. 
- **spatial correlation:** This means correlation of errors because measurements were collected over a spatial region Responses (and errors) measured close together in space are more likely to be similar in value that responses measured further apart in space. For example, home values a sample of houses in St. Paul are likely spatially correlated since the value of a house is likely more similar to it's neighbor than to a house across town. 

One method of checking for these two types of dependence is to plot the model residuals $r_i$ (y-axis) against a variable that measures, or is associated, with time or space. For the time example, plot residuals against day of the year. For the spatial example, plot residuals against a categorical "neighborhood" variable. 


### Robustness against violations

**Robustness** of a statistical method means the conclusions we make using the method aren't all that senstive to assumptions used to construct the method. Can we trust our SLR model inferences if a model assumption is violated? 

Inference about $\beta_0$, $\beta_1$, and $\mu_{y \mid x}$ from a SLR model are only robust against violations of the normality assumption. If $n$ is large enough, then the Central Limit Theorem tells us that t-tests and t-CIs are still accurate. 

SLR modeling is **not** robust against any other violations:

- **Linearity:** If your fitted mean model doesn't match the form of the true mean then you will get incorrect inferences about predictor effects, mean responses, predicted responses, etc. Even your estimated model SD will be wrong!
- **Constant variance and independence:** If violated, the SE's produced by the SLR model fit will not accuractely reflect the true uncertainty in our estimated parameters or predictions.
- **Normality for prediction** If the normality assumption is violated, then our prediction intervals will not capture the value of a new response "95% of the time". (There is no CLT and "large n" to help us here when we are trying to predict *one* response!)


### "Fixes" to violations

Here are some suggestions if a particular assumption is violated. The first part to consider "fixing" is linearity. If the mean function is not correctly specified then that will likely cause the other assumptions to not hold.

- **Linearity:** transform one or both variables to a different scale (e.g. logarithm, square root, reciprocal), modify the mean function (e.g. add a quadratic term), try non-linear regression
- **Constant variance:** tranform the response variable, weighted regression
- **Normality:** transform the response variable
- **Independence:** add more predictors, use a model with correlated errors (e.g. mixed effects, time series, spatial, etc)

--------------------------------

## Example: SLR assumptions (day 4/5) {#slr-example2}



### Drug offender sentences
The data set `DrugOffenses2.csv` contains data on 24,011 individuals convicted on federal drug charges during 2013 and 2014. We will subset these individuals to look only at cases given non-life prison sentences (`sentence2 > 0`), then look at the SLR of an individual's sentence length (`sent.nonlife`) in months against their [crimal history points](https://en.wikipedia.org/wiki/United_States_Federal_Sentencing_Guidelines#Criminal_history) (`CRIMPTS). 

#### Basic EDA shows that both sentencing and points variables are right skewed, and there are non-responses in each variable.

```{r, fig.width=5, fig.height=4, cache = TRUE}
drug <- read.csv("http://people.carleton.edu/~kstclair/data/DrugOffenses2.csv")
dim(drug)
library(dplyr)
drugPrison<- filter(drug, sentence2 > 0)
par(mfrow=c(1,2))
hist(drugPrison$CRIMPTS, main="criminal points")
hist(drugPrison$sent.nolife, main="sentence")
par(mfrow=c(1,1))
summary(drugPrison$CRIMPTS)
summary(drugPrison$sent.nolife)
```

#### Usual scatterplot
The usual scatterplot suffers from **overplotting** when we have a large data sets or datasets with very discrete variables. It is hard to see where the majority of cases are in this plot and difficult to discern any trend in the response as we change the predictor variable. 

```{r, fig.width=5, fig.height=4, cache = TRUE}
library(ggplot2)
ggplot(drugPrison, aes(x=CRIMPTS, y=sent.nolife)) + 
  geom_point() + 
  geom_smooth(method= "lm", se=FALSE)
```


#### Jittering and a *loess* smoother
Two ways to minimize the effect of overplotting are to 

- "jitter" the plotted points by adding a small amount of random noise. This can be done by using `geom_jitter` instead of `geom_point`
- use a more transparent plotting symbol. Then when points are overlapping, we will see darker colors which indicate a higher density of observations in that region of the graph. We do this by adding `alpha=` to the `geom_jitter` and make the `alpha` value less than 1 (smaller values mean more transparent points).

We will also use a `loess` smoother line to the scatterplot to help reveal the true trend in sentencing as the point history changes. A smoother is a non-parameteric way to locally "average" the response as we change the predictor value. (Non-parametric means this isn't a parameterized curve like a line or quadratic function, which means we don't end up with a nice formula for $y$ given $x$ from such a model.) We fit the `loess` model to our plot by adding another layer of `geom_smooth(method = "loess", se=FALSE)`. In the commands below, the `aes(color=)` just adds a color legend to the plot.

```{r, fig.width=5, fig.height=4, cache = TRUE}
ggplot(drugPrison, aes(x=CRIMPTS, y=sent.nolife)) + 
  geom_jitter(alpha = .2) + 
  geom_smooth(method= "lm", se=FALSE, aes(color="lm")) + 
  geom_smooth(method="loess", se=FALSE, aes(color="smoother"))
```


#### Residual plot and `loess` smoother
The scatterplot and loess smoother suggest there is slight quadratic relationship between points history and sentence length, with a positive trend up to about 20 points and a negative trend after. We can see this too in the residual plot with loess smoother added. The variation around the regression line seems fairly similar for any points value up to about 20 points, and after 20 points there are fewer cases which seem to be slightly less variable. (Note that "similar variation" means the spread around the line is of similar *magnitude for any value of x*, not that the spread around the line is *symmetric*. As we see next, the spread around the line (residuals) is skewed right.)

```{r, fig.width=5, fig.height=4, cache = TRUE}
drug.lm <- lm(sent.nolife ~ CRIMPTS, data=drugPrison)
library(broom)
drugPrison.aug <- augment(drug.lm)
ggplot(drugPrison.aug, aes(x=CRIMPTS, y=.resid)) + 
  geom_jitter(alpha = .2) + 
  geom_smooth(method= "lm", se=FALSE, aes(color="lm")) + 
  geom_smooth(method="loess", se=FALSE, aes(color="smoother"))
```

#### Checking normality
The residuals for this model are not normally distributed, there is a strong skew to the right. **If** the linear model fit the trend of the data, this non-normality would **not** be a concern for inference about the mean parameters $\beta$. It would only be a concern if we wanted to **predict** one individual's sentence length with confidence given their criminal history points. 

```{r, fig.width=5, fig.height=4, cache = TRUE}
par(mfrow=c(1,2))
hist(resid(drug.lm), main="Histogram of residuals")
plot(drug.lm, which = 2)
```


### Case study 15.2 - Global Warming
Let's look at the regression of global mean temperature deviation (Celsius) on year (1850 through 2010). The scatterplot suggests a curved relationship with time. Temperature deviation is the mean yearly temp minus the mean temp from all 161 years.

```{r}
library(Sleuth3)
temps <- case1502
summary(temps)
ggplot(temps, aes(x = Year, y = Temperature)) + 
        geom_point() + 
        geom_smooth(method = "lm", se = FALSE) + 
        geom_smooth(method = "loess", se = FALSE, linetype = "dashed")
```

#### Independent residuals?
After fitting our regression of `Temperature` on `Year` and `Year^2` (due to curvature), a check of the residual plot (residuals vs. `Year`) shows no obvious signs of curvature in residuals so the quadratic mean function with time seems adequate.

```{r}
temps.lm <- lm(Temperature ~ Year + I(Year^2), data=temps)
temps.aug <- augment(temps.lm)
ggplot(temps.aug, aes(x = Year, y = .resid)) + 
        geom_point() + 
        geom_hline(yintercept = 0, linetype = "dashed")
```

The independence assumption will be violated for this model if residuals closer in time (x) are more similar than residuals further apart in time. This is hard to see in the basic residual plot above. To see this idea better, we can look at line plot of residuals vs. time:

```{r}
ggplot(temps.aug, aes(x = Year, y = .resid)) + 
        geom_point() + 
        geom_line() + 
        geom_hline(yintercept = 0, linetype = "dashed")
```

With this plot is it clear that if one year has a negative residual (lower than expected temps), the next year's residual is often also negative. Same is true for positive residuals. This indicates that there is a temporal association in these residuals, which means that they are not indepedent. A SLR model is not appropriate for this data because the SEs will not accurately reflect the uncertainty in our estimates when independence is violated. 

#### Autocorrelation
The **correlation** statistic is used to assess the strength and direction of a linear relationship between two quantitative variables. The **autocorrelation** is used to assess how correlated values are for the **same** quantitative variable. Autocorrelation is computed by pairing responses by a **lag** amount. Lag of 1 means we are computing correlation for responses that differ by one spot in the vector of responses. For our time and temp data, this means looking at the correlation between responses that differ by a year. Lag 2 means looking at correlation between responses that differ by 2 years, and so on. 

We can use autocorrelation to verify that residuals closer together in time are more similar than those further apart in time:

```{r}
acf(resid(temps.lm))
acf(resid(temps.lm), plot=FALSE)
```

Residuals that differ by one year have a correlation of 0.592, indicating a moderate positive correlation. This is what were able to see in the line plot above. Residuals that differ from 2-9 years have correlations between about 0.45 to 0.25. After  years there is little to no autocorrelation between residuals. 


## Transformations {#slr-trans}

Transformations of one or both variables in SLR is done to fix one, or both, of these assumption violations: linearity and constant variance. Tranformations are often explored via trial-and-error, use plots of transformed variables and residual plots from potential transformed SLR models to find the "best" SLR model. "Best" is determined by a model that **most satisfies the SLR assumptions**, not a model that yields the "strongest" relationship between variables. A transformation choice can also be determined from a theoretical model that we have from a specific application.

### Transformation choices
We need to look at non-linear functions of a variable. A linear transformation, like changing a measure from cm to feet would only change the numbers you see on the axes label and not the relative location of points on the plot. Common transformations to consider are the following:

- **logarithms**: Any log-base can be explored but base choice will only affect how you interpret a model, not affect the linearity or constant variance assumptions. E.g. if $\log_2(y)$ vs. $x$ looks nonlinear, then $\log_{10}(y)$ vs. $x$ will also be nonlinear. Logarithms are nice because they are easily interpretable in a SLR model (more to come). In R, 
```{r}
log(2)  # the natural log function (ln)
log10(2) # base 10
log2(2)  # base 2
```
- **square root**: A square root transformation is an "in between" no transformation and a logarithm transformation (see the Figure below) 
```{r}
sqrt(4)  # square root function
```

```{r, echo=FALSE}
ggplot(data.frame(x=c(.01, 10)), aes(x=x)) + 
  stat_function(fun = sqrt, aes(color="square root")) + 
  stat_function(fun = log, aes(color="logarithm (ln)")) + 
  geom_line(aes(y=x, color="linear")) + 
  labs(x="variable", y="transformation of the variable", color="function", 
       title="common transformations")
```

### Transformations in R
We can visualize square root and logarithms in a`ggplot2` scatterplot by adding a layer that changes the scale of an axes. For example, `scale_x_sqrt()` will change the `geom_point()` scatterplot to a plot of $y$ aganist $\sqrt{x}$ but the axes numeric labels on the x-axis will still measure $x$. The layer `scale_y_log10()` will convert $y$ to the base-10 logarithm. If you want a transformation that is not one of these two, then you can transform the variable in the aesthetic: `ggplot(data, aes(x = 1/x, y = y))` would plot $y$ against the reciprocal $1/x$. 

You can apply transformations directly in the `lm` command too, e.g. `lm(log(y) ~ log(x), data)` fits the regression of $\log(y)$ against $\log(x)$. If a transformation involves a mathematics function (like power `^` or division `/`) on the right side of the formula symbol `~`, then you need to use the "as is" operator function `I()`. E.g. the regression of $y$ against the reciprocal $1/x$ is `lm(y ~ I(1/x), data)`. 

### Interpretation
In general, models with transformed variables should be interpreted as a usual SLR but with the transformed variable scale. For example, the SLR for $\sqrt{y}$ against $\sqrt{x}$ has a mean function that looks like
$$
\mu_{\sqrt{y} \mid x} = \beta_0  + \beta_1 \sqrt{x}
$$
so a one unit increase in the **square root** of $x$ in associated with a $\beta_1$ change in the mean of the **square root** of $y$. 

Models that use logarithms have a nicer interpretation. Section \@ref(slr-logs) has more review of logarithms but the three common SLR models that use logarithms are given here. 

- **Logarithmic model**: The regression of $Y$ on $\log(x)$ has the mean function
$$
\mu(Y \mid \log(x)) = \beta_0 + \beta_1 \log(x)
$$
Multiplying $x$ by a factor $m$ is associated with a mean function change of 
$$
\mu(Y \mid \log(mx)) - \mu(Y \mid \log(x)) = \beta_1(\log(m) + \log(x)) - \beta_1 \log(x) = \beta_1 \log(m)
$$
If **base-2** is used for the tranformation of $x$ then a **doubling** of $x$ (so $m=2$) is associated with a  change in mean $y$ of $\beta_1 \log_2(2) = \beta_1$.
- **Exponential model**: The regression of $\log(Y)$ on $x$ has the mean function
$$
\mu(\log(Y) \mid x) = median(\log(Y) \mid x) = \beta_0 + \beta_1 x
$$
Since the median of a logged-variable equals the log of the median of the variable, we can "untransform" this model and write it in terms of the median of $Y$ (assuming here that `log` is the natural log):
$$
median(Y \mid x) = e^{\log(median(Y \mid x))} = e^{median(\log(Y) \mid x)} = e^{\beta_0}e^{\beta_1 x}
$$
Note that the *mean* of a logged-variable *does not* equal the log of the mean of the variable, so we can't express the untransformed model in terms of the median of $Y$.

A one unit increase in $x$  is associated with a $e^{\beta_1}$-factor change in the *median* function since
$$
median(Y \mid x+1) = e^{\beta_0}e^{\beta_1 (x+1)} = e^{\beta_0}e^{\beta_1 x}e^{\beta_1} = median(Y \mid x) e^{\beta_1} 
$$
- - **Power model**: The regression of $\log(Y)$ on $\log(x)$ has the mean function
$$
\mu(\log(Y) \mid \log(x)) = median(\log(Y) \mid\log(x)) = \beta_0 + \beta_1 \log(x)
$$
Using the same logic as the exponential model, inference on the untransformed scale of $Y$ is about the *median* of $Y$:
$$
median(Y \mid x) = e^{\log(median(Y \mid x))} = e^{median(\log(Y) \mid x)} = e^{\beta_0}e^{\beta_1 \log(x)} = e^{\beta_0}(e^{\log(x)})^{\beta_1} =  e^{\beta_0}x^{\beta_1}
$$

An m-fold (multiplicative) change in $x$  is associated with a $m^{\beta_1}$-factor change in the *median* function since
$$
median(Y \mid mx) = e^{\beta_0}e^{\beta_1 \log(mx)} = e^{\beta_0}(e^{\log(x)})^{\beta_1}(e^{\log(m)})^{\beta_1}  = median(Y \mid x) m^{\beta_1} 
$$

### Review: Logarithms {#slr-logs}

Let $b>0$ and $x>0$. The logarithm (base-$b$) of $x$ is denoted $\log_b(x)$ and equal to 
$$
\log_b(x) = a
$$
where $a$ tells us what power we must raise $b$ to to obtain the value $x$:
$$
b^a = x
$$

Easy examples are: $b=2$, $x=8$ and $a=3$,
$$
\log_2(8) = 3
$$
since $2^3 = 8$. Or using base $b=10$, then
$$
\log_{10}(0.01) = -2
$$
since $10^{-2} = 0.01$. 

Some basic facts logarithm facts are
$$
\log_b(b) = 1
$$
since $b^1 = b$ and 
$$
\log_b(1) = 0
$$
since $b^0 = 1$. 

#### Interpreting logged variables
**Multiplicative** changes in $x$ result in **additive** changes in $\log_b(x)$. If $m>0$, then 
$$
\log_b(mx) = \log_b(m)  + \log_b(x) 
$$
For example, 
$$
\log_2(16) = \log_2(2\times 8) = \log_2(2) + \log_2(8) = 1 + 3 = 4
$$


#### Inverse (i.e. reversing the log, getting rid of the log, ...)
The logarithm and exponential functions are inverses of one another. This means we can "get rid" of the log by calculating $b$ raised to the logged-function:
$$
b^{\log_b(x)} = x
$$
This will be useful in regression when we have a linear relationship between logged-response $y$ and a set of predictors. We need to 

For example, suppose we know that 
$$
\log_2(y) = 3 + 5x
$$
To return this to an expression on the original (unlogged) scale of $y$, we need take both sides raised to the base 2:
$$
2^{\log_2(y)} = 2^{3 + 5x}
$$
Simplifying both sides gives
$$
y = 2^3 \times 2^{5x}
$$



#### Questions
Solutions are posted on the class Moodle site.

1. Write the following as the sum of two logarithms. Simplify as much as possible:
 
    a. $\log_2(2x)$
    b. $\log_2(0.5x)$
    c. $\ln(2x)$ where $\ln$ is the natural log (base-$e$)

2. Write the following expressions in terms of $y$, not $\log(y)$. Simplify as much as possible:

    a. $\log_2(y) = 1 - 3x$
    b. $\log_{10}(y) = -2 + 0.4x$
    c. $\ln(y) = 1 - 3x$
    
3. Write the following expressions in terms of $y$ and $x$, not $\log(y)$ and $\log(x)$. Simplify as much as possible:

    a. $\log_2(y) = 1 - 3\log_2(x)$
    b. $\ln(y) = -2 + 0.4\ln(x)$
    c. $\ln(y) = 1 - 3\log_2(x)$
    
4. Logarithmic model: Regression of $Y$ on $\log(x)$ obtains the following estimated mean of $Y$:
$$
\hat{\mu}(Y \mid x) = 1 - 3 \log_2(x)
$$

    a. What is the  change in estimated mean response if we double the value of $x$?
    b. What is the  change in estimated mean response if we triple the value of $x$?
    c. What is the  change in estimated mean response if we reduce the value of $x$ by 20%?


5. Exponential model: Regression of $\log_2(Y)$ on $x$ obtains the following estimated *median* of $Y$:
$$
\hat{median}(\log_2(Y) \mid x) = -2 + 0.4x
$$

    a. Write the median in terms of $Y$ instead of $\log_2(Y)$. Simplify as much as possible. 
    b. What is the multiplicative change in estimated median response if we increase $x$ by 1 unit?
    c. What is the percent change in estimated median response if we increase $x$ by 1 unit?
    d. What is the multiplicative change in estimated median response if we decrease $x$ by 2 units?
    e. What is the percent change in estimated median response if we decrease $x$ by 2 units?

6. Power model: Regression of $\log_2(Y)$ on $\log_2(x)$ obtains the following estimated *median* of $Y$:
$$
\hat{median}(\log_2(Y) \mid x) = 1 -3\log_2(x)
$$

    a. Write the median in terms of $Y$  and $x$ instead of $\log$s. Simplify as much as possible. 
    b. What is the multiplicative change in estimated median response if we increase $x$ by 50%?
    c. What is the percent change in estimated median response if we increase $x$ by 50%?
    d. What is the multiplicative change in estimated median response if we reduce the value of $x$ by 20%?
    e. What is the percent change in estimated median response if we reduce the value of $x$ by 20%?


## Example: Transformations
