# Simple Linear Regression {#slr}

This chapter contains content from Sleuth Chapters 7 and 8. 

## The variables
Suppose we have a quantitative response variable $y$ that we want to relate to an explantory (aka predictor) variable $x$. For now, we will assume that $x$ is also quantitative. 


## The model form {#slr-model}
This section describes the SLR model for a particular **population** of interest. Another way to frame the model is that it describes a hypothetical **data generating process (DGP)** that was used to generate the sample of data that we have on hand. 

Let $Y_i$ be the response from unit $i$ that has explanatory (aka predictor) value $x_i$. There are two **equivalent** ways to express the SLR model for $Y$:

- Conditional normal model: Given a value $x_i$, the response $Y_i$ follows a normal model with mean and SD given below:
$$
Y_i \mid x_i \sim N(\mu_{y\mid x} = \beta_0 + \beta_1 x_i, \sigma)
$$

- Mean + error: Statisticians are more likely to use a model specification that expresses $Y$ as a function of the **expected value/mean** of $Y$ plus an **error** term that models variation in responses around the mean:
\begin{equation}
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i \ \ \ \ \ \epsilon_i \sim N(0, \sigma)
\end{equation}

**Both** expressions of the SLR model above say the same thing: 

- **Linear Mean:** $\mu_{y\mid x} = E(Y \mid x) =  \beta_0 + \beta_1 x$ describes the population mean value of $Y$ given a predictor value $x$. This mean value varies linearly with $x$ and the population parameters are $\beta_0$ and $\beta_1$. 
- **Constant SD:** $SD(Y\mid x)=\sigma$ describes the SD of $Y$'s in the population around a given mean value $\mu_{y\mid x}$. The fact that this SD **does not** depend on the value of $x$ is called the contant variance, or homoscedastic, assumption. 
- **Normality:** The shape of population response values around $\mu_{y\mid x}$ is described by a normal distribution model.  

Finally, one last assumption is made for the SLR model:

- **Indepedence:** Given a predictor value of $x$, all responses $Y$ occur independently of each other. 

There are a total of **three parameters** in the SLR model: 

- the two mean parameters $\beta_0$ and $\beta_1$
- the SD parameter $\sigma$

### Interpretation

For a SLR model:

- $\beta_0$ is the mean response when the predictor value is 0 since $\mu_{y \mid 0} = \beta_0 + \beta_1(0) = \beta_0$.
- $\beta_1$ tells us how the mean response changes for a one unit increase in $x$

### Example: Woodpecker nests {#slr-model-ex}

We want to model nest depth (cm) in a tree cavity as a function of ambient air temperature (Celsius). Our SLR model for this relationships says that, given an ambient air temp $x_i$, a randomly selected nest will have a depth $Y_i$ that is modeled as
$$
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i \ \ \ \ \ \epsilon_i \sim N(0, \sigma)
$$
This means that depth is normally distributed with a mean of $\mu_{depth \mid temp} = \beta_0 + \beta_1 (temp)$ and a SD of $\sigma$. 

-----------

## Theory: Estimation {#slr-est}

Let's consider taking a random sample of size $n$ of responses and predictor values from our population (or DGP) for which the SLR model holds: $(x_1, Y_1), \dotsc, (x_n, Y_n)$. The notation here, $(x_i, Y_i)$ implies that the predictor value $x_i$ is fixed, but $Y_i$ is a **random variable** that is generated from the SLR model described in Section \@ref(slr-model). The **estimation** problem is that we need to use the sample of size $n$ to estimate the SLR model parameters $\beta_0$, $\beta_1$ and $\sigma$.

Once we observe a sample of size $n$, then we can use the SLR model to determine the probability of observing the sample. This probability, which depends on the actual model parameters $\beta_0$, $\beta_1$ and $\sigma$, is called a **likelihood function**. We plug the observed data into this function, then find the parameters values that **maximize** the function using calculus. For a SLR model, this process yields the following **maximum likelihood estimators (MLE)** of our parameters: 
$$
\hat{\beta}_1 = \dfrac{\sum_{i=1}^n (x_i - \bar{x})(Y_i - \bar{Y})}{\sum_{i=1}^n (x_i - \bar{x})^2} \ \ \ \ \ \hat{\beta}_0 = \bar{Y} - \bar{\beta}_1 \bar{x} \ \ \ \ \ \hat{\sigma} = \sqrt{\dfrac{\sum_{i=1}^n (Y_i - \hat{y}_i)^2}{n-2}}
$$
where $\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$ is the **predicted value** of $y$ given the value $x_i$.

### Sampling Distributions for SLR estimates

The sampling distribution of a model estimate ($\hat{\beta}_0$ or $\hat{\beta}_1$) is constructed by:

1.  fix a set of predictor values: $x_1, \dotsc, x_n$
2.  for each fixed $x_i$, generate a response $Y_i$ from $N(\beta_0 + \beta_1 x_i, \sigma)$
3. compute the MLE's $\hat{\beta}_0$ and $\hat{\beta}_1$ from the observed sample from (2)
4. repeat 2-3 lots of times, then the distribution of the estimates from part (3) show the sampling distribution of the slope or intercept estimate.

Using probability theory, we can show that the sampling distributions of both $\hat{\beta}_0$ and $\hat{\beta}_1$ are approximately normal when $n$ is "large enough" (thanks to the CLT):
$$
\hat{\beta}_i \sim N(\beta_i, SD(\hat{\beta}_i))
$$

- **Unbiased:** We see that the expected value (mean) of $\hat{\beta}_i$ is the parameter $\beta_i$, meaning it is an unbiased estimator. (It doesn't systematically over- or under-estimate the parameter of interest.)
- **Standard error:** We end up estimating the SD in the sampling distribution given above. The SEs for each mean parameter estimate are
$$
SE(\hat{\beta}_1) = \hat{\sigma}\sqrt{\dfrac{1}{(n-1)s^2_x}} \ \ \ \ \ SE(\hat{\beta}_0) = \hat{\sigma}\sqrt{\dfrac{1}{n} + \dfrac{\bar{x}^2}{(n-1)s^2_x}}
$$

-----------

## SLR model simulation {#slr-sim}

Download the Markdown of this activity: [.Rmd](https://kstclair.github.io/Math245/Rhandouts/day3RegLineSim.Rmd).


### Simulation function
This chunk contains code that defines our function `reg.sim` that **simulates $n$ responses from a given regression model and given set of $n$ predictor values $x$**. I've excluded it from our compiled document so see the .Rmd file to take a look at how this was created. 

```{r, include=FALSE}
reg.sim <- function(x, beta0, beta1, sigma, grph=T)
{
  # function takes a sample of responses y generated from the given model/x values
  # returns estimated intercept and slope
  # x = explantory variable
  # beta0, beta1 = true model intercept and slope
  # sigma = true model SD
  # grph = print simulation graphs (y vs. x)? (default is true)
  n<- length(x)
  muy.x <- beta0 + beta1 * x

  # simulate some y's, then fit lm using these y's:
  y <- rnorm(n, muy.x,sigma)
  mylm <- lm(y~x)

  # plot data, LS line, true line
  if (grph){
    plot(x,y,xlab="x",ylab="simulated y values",pch=16, ylim=c(0,30))
    legend("topright",legend=c("Population line", "Sample line"),col=c("red","black"),lty=c(1,1),lwd=2)
    abline(mylm,lwd=2)
    abline(beta0,beta1,lwd=2,col="red")
    }
  return(list(b0=mylm$coefficients[[1]], b1=mylm$coefficients[[2]] ))
}
```

### Run the function once
Let's use the function from (1) above. We will use the $n=12$ temps (x-values) from the woodpeckers data and assume that the **true** model is:
$$\mu(y \mid x) = 20 - 0.4x$$ 
(red line below) with  $\beta_0=20$, $\beta_1=-0.4$, and $\sigma=2$. In the code below I use the `set.seed()` command to "fix" the random number generator so I get the same answer each time this is run (so my answer in the handout is reproduced each time this file is compiled).

```{r, fig.width=5,fig.height=3}
wpdata<- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
set.seed(77)  
reg.sim(x=wpdata$temp, beta0=20, beta1=-.4, sigma=2)
```

For this simulated sample (with seed of 77), the estimated regression line is $\hat{\mu}(y \mid x) = 20.889 - 0.468x$ (black line).

### Simulated sampling distribution for $\hat{\beta}_1$
We will now use the `replicate` command to generate 1000 different samples which create 1000 different estimates of $\beta_1$. A histogram of these estimates simulates the sampling distribution of estimated slope.  

**What is the shape of the sampling distribution? Where is the distribution centered? How variable are these estimated slopes.**

```{r, fig.width=5,fig.height=3.5 }
set.seed(7)  # just makes simulation reproducible
slopes<- replicate(1000,reg.sim(x=wpdata$temp, beta0=20, beta1=-.4, sigma=2,grph=F)$b1)
hist(slopes); abline(v=-0.4,col="red", lwd=2)
mean(slopes); sd(slopes)
```


### Are slope and intercept estimates correlated?
In regression, it is not unusual to be interested in estimating a **linear combination** of our model parameters. An easy example of such a combination of parameters is the mean response for a given value $x_0$ of the predictor: 
$$
\mu(y \mid x=x_0) = \beta_0 + \beta_1x_0
$$
For a specific example, we may want to estimate the mean response (depth) for a temp of $x_0=5$ degrees: $\mu(y \mid x=5) = \beta_0 + \beta_1 5$. If we don't know $\beta_0$ and $\beta_1$, then this mean parameter is a linear combination of two unknown parameters which we need to estimate. 

The natural estimate of this is just the estimated mean response: $\hat{\mu}(y \mid x=x_0)=\hat{\beta}_0 + \hat{\beta}_1x_0$. To assess how variable this estimate is (i.e. to get its SE) we need to understand how (if) the individual estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ are correlated. We can use a simulation to look at this issue by generating 1000 samples from our model and plotting each $(\hat{\beta}_0,\hat{\beta}_1)$ pair for each sample. 

**How are estimated slope and intercept associated? Any ideas why?**

```{r}
set.seed(7)  # this seed MUST match the seed used to get slopes!
intercepts<- replicate(1000,reg.sim(x=wpdata$temp, beta0=20, beta1=-.4, sigma=2,grph=F)$b0)
plot(intercepts,slopes); abline(h=-.4,col="red"); abline(v=20,col="red")
title("Estimated slopes and intercepts")
cor(intercepts, slopes) # correlation between estimates
cov(intercepts, slopes) # covariance between estimates
```




--------------------------------

## Inference for mean parameters {#slr-inf}

In intro stats, you used $t$ inference procedures for inference about population means since: (1) the sampling distribution of the sample mean was normally distributed and (2) we had to estimate its variability with a SE. The same goes for inference about the mean response parameters in a SLR model:
$$
t = \dfrac{\hat{\beta}_i - \beta_i}{SE(\hat{\beta}_i)} \sim t_{df=n-2}
$$
Use a t-distribution with $n-2$ degrees of freedom for inference about the mean parameters. The degrees of freedom are calculated as the sample size $n$ minus the number of terms in $\mu_{y \mid x}$ that you have to estimate with the data. 


### Confidence Intervals
To estimate either mean parameter with $C$% confidence, we have the general form
$$
\hat{\beta}_i \pm t^* SE(\hat{\beta}_i)
$$
where $t^*$ is the $(100-C)/2$ percentile from the t-distribution with $df=n-2$ degrees of freedom. 

### Hypothesis tests
We can test the hypothesis
$$
H_0: \beta_i = \beta^*_i
$$
with the following t-test statistic:
$$
t =\dfrac{\hat{\beta}_i - \beta^*_i}{SE(\hat{\beta}_i)}
$$
where $\beta^*$ is our hypothesized value of $\beta$ (intercept or slope). The t-distribution with $n-2$ degrees of freedom is used to compute the p-value that is appropriate for whatever $H_A$ is specified. 

The usual test results given by standard regression output tests whether a parameter value (intercept or slope) is equal to 0 vs. not equal to 0:
$$
H_0: \beta_i = 0 \ \ \ \ \ H_A: \beta_i \neq 0
$$
with a test stat of 
$$
t =\dfrac{\hat{\beta}_i - 0}{SE(\hat{\beta}_i)}
$$

----------------------------

## Inference for average or predicted response {#slr-inf2}

### Confidence intervals for $\mu_{y \mid x}$

Here we are interested in estimating not just a $\beta$ with confidence, but we want to estimate the **mean** response for a given value of $x_0$. For example, suppose we want to estimate the **mean** nest depth when the temp is $x_0=8$ degrees. This mean parameter of interest is then:
$$
\mu_{depth \mid temp=8} = \beta_0 + \beta_1 (8)
$$


Our parameter of interest is $\mu_{y \mid x_0} = \beta_0 + \beta_1 x_0$ where $x_0$ is known and $\beta$'s need to be estimated. The natural estimator is just the fitted equation:
$$
\hat{\mu}_{y \mid x_0} = \hat{\beta}_0 + \hat{\beta}_1 x_0
$$
As will *any* estimator, we can measure the variability of this estimator with a SE:
$$
SE(\hat{\mu}_{y \mid x_0}) = \hat{\sigma} \sqrt{\dfrac{1}{n} + \dfrac{(x_0 - \bar{x})^2}{(n-1)s^2_x}}
$$
**Note!** This SE **depends on $\pmb{x_0}$**! It is miminized when $x_0$ equals the mean predictor value $\bar{x}$ and it grows as $x_0$ gets further from $\bar{x}$. Estimation is most precise in the "middle" of the predictor range and becomes less precise at the "edges" (where we usually have less data). 

A 95% confidence interval for the mean response $\mu_{y \mid x_0}$ looks like
$$
\hat{\mu}_{y \mid x_0} \pm t^*_{df=n-2}SE(\hat{\mu}_{y \mid x_0}) 
$$

### Prediction intervals for new cases
To predict *one individual's future response* $Y$ for the predictor value $x_0$, we just use the  fitted equation:
$$
pred_{y \mid x_0} = \hat{\beta}_0 + \hat{\beta}_1 x_0
$$
(We use $pred_{y \mid x_0}$ to remind us which predictor value as used for prediction.) 

Recall that the SLR model assumes the $Y$ values when $x=x_0$ are normally distributed with mean $\mu_{y \mid x_0}$ and SD $\sigma$. The SE of this prediction at $x_0$ takes into account (1) uncertainty in estimating the *mean*  $\mu_{y \mid x_0}$  and (2) variation in $Y$'s around the mean response ($\sigma$):
$$
SE(pred_{y \mid x_0}) = \hat{\sigma} \sqrt{1 + \dfrac{1}{n} + \dfrac{(x_0 - \bar{x})^2}{(n-1)s^2_x}} = \sqrt{\hat{\sigma}^2 + SE(\hat{\mu}_{y \mid x_0})^2  }
$$
A 95% **prediction** interval for a future individual response at $x=x_0$ looks like
$$
pred_{y \mid x_0} \pm t^*_{df=n-2}SE(pred_{y \mid x_0})
$$

Predictions intervals feel and look similar to the mean response intervals above, but there is a very important conceptual difference: *prediction* means we are trying to "guess" at *one individual response* as opposed to the mean response of a large group of individuals. For example, if we want to understand nest depths for all nest build when temp is 8 degrees, then we care about estimating a fixed (but unknown) mean depth $\mu_{depth \mid temp=8}$. If we see a bird starting to build a nest at 8 degrees, then we care about predicting this one, randomly determined depth $Y$ using $pred_{depth \mid temp=8}$ and would use a prediction interval. 


--------------------------------

## Example: SLR model (day 3) {#slr-example1}
We will revisit the woodpecker nesting data first described in Section \@ref(slr-model-ex). 

### Load data    
Let's suppose these 16 nests are a random sample of all nests in the collection region. The command `head(dataname)` produces a view of the first 5 rows of data.

```{r, eval = FALSE}
wpdata <- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
head(wpdata)
dim(wpdata)
```
```{r, include = FALSE}
wpdata <- read.csv("http://people.carleton.edu/~kstclair/data/woodpeckers.csv")
```

### EDA
Start with univariate exploration:

```{r, eval = FALSE}
summary(wpdata)
par(mfrow=c(1,2))
hist(wpdata$temp)
hist(wpdata$depth)
par(mfrow=c(1,1))  
```

Graphically explore the ("bivariate") relationship between temp and depth with a scatterplot using the command `plot(y,x)`. Here is the `plot` version (with `pch` point character changed to give filled circles):

```{r, fig.width=6, fig.height=4, eval = FALSE}
plot(depth ~ temp, data=wpdata, pch=16, xlab="air temperature (C)", 
     ylab="nest depth (cm)", main="woodpeckers scatterplot")
```

Here is the `ggplot2` version with `labs` added to change labels and title:
```{r, fig.width=6, fig.height=4, eval = FALSE}
library(ggplot2)
ggplot(wpdata, aes(x=temp, y = depth)) + geom_point() +
  labs(x = "air temperature (C)", y= "nest depth (cm)", title= "woodpeckers scatterplot")
```

### The least squares line (the estimated SLR model):
You fit the linear model with mean $\mu_{Y \mid x} = \beta_0 + \beta_1 x$ with the linear model function `lm(y ~ x, data=)`.

```{r, eval = FALSE}
wood.lm <- lm(depth~temp, data=wpdata)
wood.lm
```
```{r, include = FALSE}
wood.lm <- lm(depth~temp, data=wpdata)
wood.lm
```
We have mean parameter estimates:
$$
\hat{\beta}_0 = `r round(wood.lm$coefficients[1],3)` \ \ \ \ \hat{\beta}_0 = `r round(wood.lm$coefficients[2],3)`
$$

The object `wood.lm` is called a *linear model object* in R. We can add the regression line from this object to an existing *base R* plot of the data using the `abline` command.
```{r, fig.width=6, fig.height=4, eval = FALSE}
plot(depth ~ temp, data=wpdata, pch=16, xlab="air temperature (C)", 
     ylab="nest depth (cm)", main="regression of depth on temp")
abline(wood.lm)
```

The `ggplot2` package contains a `geom_smooth` geometry to add this SLR line:
```{r, fig.width=6, fig.height=4, eval = FALSE}
ggplot(wpdata, aes(x=temp, y = depth)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "air temperature (C)", y= "nest depth (cm)", title= "woodpeckers scatterplot")
```

With `geom_smooth`, you need to specify the type of method used to relate $x$ to $y$. Adding `se=FALSE` removes a confidence interval for the mean trend. Notice that the `aes` aesthetic given in the `ggplot` function specifies the x and y variables to plot. The `geom`'s that follow this part of the command use this `aes` to create points and an estimated line.   



### Inference for coefficients
The `summary` command is used for statistical inference for the intercept $\beta_0$ and slope $\beta_1$:

```{r, eval = FALSE}
summary(wood.lm)
```
This output gives the  mean parameter estimates in the `Estimate` column of the main table:
$$
\hat{\beta}_0 = `r round(wood.lm$coefficients[1],3)` \ \ \ \ \hat{\beta}_0 = `r round(wood.lm$coefficients[2],3)`
$$
The estimated model SD $\sigma$ is given by `Residual standard error`:
$$
\hat{\sigma} = `r round(summary(wood.lm)$sigma,3)`
$$


You should know how to verify (or find) the test stats and p-values for $\beta_0$ and $\beta_1$ given by the `summary` command if you are given the estimates and SEs. For the slope test $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$ we get a  $t$ test stat of

```{r, eval = FALSE}
(-0.34218 - 0)/0.05961
```

The p-value for this two sided test is the probability of being above +5.741 and below -5.741, or double the probability below -5.741 (since the t-distribution is symmetric around 0):

```{r, eval = FALSE}
pt(-5.741,10)
2*pt(-5.741,10)
```

A 95% confidence interval for the slope $\beta_1$ is computed from the t distribution with 10 degrees of freedom:

```{r, eval = FALSE}
qt(.975,10)
-.34218 + c(-1,1)*qt(.975,10)*.05961
confint(wood.lm)
```

The `confint` function gives the most accurate interval (no rounding error) but you need to know how to compute these CIs ``by hand."


### Additional `lm` information
The function `lm` creates a *linear model* object in R that has lots of information associated with it. Information includes the coefficient values, fitted values ($\hat{y}_i$), and residuals ($y_i - \hat{y}_i$):

```{r, eval = FALSE}
attributes(wood.lm)
wood.lm$coefficients
wood.lm$fitted.values   # predicted y values (y-hat)
wood.lm$residuals       # residuals for each data point
```

There are also functions that act on `lm` objects like
```{r, eval = FALSE}
fitted(wood.lm)
resid(wood.lm)
vcov(wood.lm)  # variance and covariance matrix of beta estimates
```


### `broom` package: Tidy `lm` output
The `broom` package contains functions that convert model objects (like a `lm` object) into "tidy" data frames. The `tidy` command summarizes model results:
```{r, eval = FALSE}
library(broom)
tidy(wood.lm)
```
Notice that the output object is called a "tibble" which is a type of data frame in R. 

Here we can add confidence intervals for the model parameters to the output:
```{r, eval = FALSE}
tidy(wood.lm, conf.int=TRUE)
```

The `augment` command augments the data frame used to create a `lm` with predicted values and residuals from the `lm` model:
```{r, eval = FALSE}
wpdata.aug <- augment(wood.lm)
head(wpdata.aug)
```
The `.fitted` column gives the fitted values $\hat{y}_i$ for each case and the `.resid` gives the residuals $y_i - \hat{y}_i$.


### Inference for the mean and predicted response
We can get a 95% confidence interval for the mean nest depth at 8 degrees Celsius $\mu(depth\mid temp=8)$  with the `predict` command with interval type `confidence` specified:

```{r, eval = FALSE}
predict(wood.lm, newdata = data.frame(temp=8), interval = "confidence", se.fit=T)
```
The argument `se.fit=T` provides the SE of the mean response estimate. The estimated mean depth of all nests built at 8 degrees is 17.38 cm with a SE of 0.697 cm. We are 95% confident that the mean depth at 8 degrees is between 15.83 to 18.9 cm. 

You can verify the computation of the SE and CI using summary stats for `temp` and the estimated parameter values from the `summary` command:
$$
\hat{\mu}_{temp \mid 8} = \hat{\beta}_0 + \hat{\beta}_1 (8) = `r round(wood.lm$coefficients[1],3)` + (`r round(wood.lm$coefficients[2],3)` )(8) = `r round(wood.lm$coefficients[1] + wood.lm$coefficients[2]*8,3)`
$$
$$
SE(\hat{\mu}_{depth \mid 8}) = \hat{\sigma} \sqrt{\dfrac{1}{n} + \dfrac{(x_0 - \bar{x})^2}{(n-1)s^2_x}} = `r round(summary(wood.lm)$sigma, 3)` \sqrt{\dfrac{1}{12} + \dfrac{(8-`r round(mean(wpdata$temp),3)`)^2}{(12-1)`r round(sd(wpdata$temp),3)`^2}} = 0.6972
$$

```{r, eval = FALSE}
nrow(wpdata)
mean(wpdata$temp)
sd(wpdata$temp)
mn.est.se <- 2.33453*sqrt(1/12 + (8-11)^2/((12-1)*11.80909^2))
mn.est.se
mn.est <- 20.12228 -0.34218*8
mn.est
mn.est + c(-1,1)*qt(.975,10)*mn.est.se
```

You can also include more than one predictor value temp in this function:
```{r, eval = FALSE}
predict(wood.lm, newdata = data.frame(temp=c(8,20)), interval = "confidence", se.fit=T)
```

We can get a 95% prediction interval for the depth of one depth constructed at 8 degrees Celsius $pred_{depth \mid temp=8}$  with the `predict` command with interval type `prediction` specified:

```{r, eval = FALSE}
predict(wood.lm, newdata = data.frame(temp=8), interval = "prediction")
```
We are 95% confident that a new nest built at 8 degrees will have a depth between 11.96 to 22.81 cm. 


R does not give us the SE for prediction $SE(pred(Y \mid x_0)) = \sqrt{\hat{\sigma}^2 + SE(\hat{\mu})^2}$ so we need to compute it by hand if we want its value:
$$
SE(pred_{depth \mid 8}) = \sqrt{\hat{\sigma}^2 + SE(\hat{\mu})^2} =  \sqrt{`r round(summary(wood.lm)$sigma, 3)`^2 + 0.6972^2} = 2.436
$$

```{r, eval = FALSE}
se.pred <- sqrt(2.33453^2 + 0.6972406^2)
se.pred
```

The predicted depth of one nest build at 8 degress is 17.38 cm with a SE of 2.44 cm. The 95% prediction interval produced above can be verified as follows:

```{r, eval = FALSE}
yhat <-  20.12228 -0.34218*8
yhat
yhat + c(-1,1)*qt(.975,10)*se.pred
```


### Adding confidence bands to a scatterplot
The `geom_smooth` function in `ggplot2` adds a 95% confidence interval for $\mu_{y \mid x}$ around the estimated mean line:
```{r, eval = FALSE}
ggplot(wpdata, aes(x=temp, y = depth)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(x = "air temperature (C)", y= "nest depth (cm)", title= "woodpeckers scatterplot with mean confidence interval")
```

Adding prediction interval bands takes slightly more work. First, create a new version of the data set that includes prediction intervals for each case in the data:
```{r, eval = FALSE}
wpdata.pred <- data.frame(wpdata, predict(wood.lm, interval="prediction"))
head(wpdata.pred)
```

Then add a `geom_ribbon` layer to the previous plot, with `ymin` and `ymax` determined by the prediction interval's lower (`lwr`) and upper (`upr`) bounds. The `fill` arguments in both layers below are not really needed, but they are used here to provide a legend label for the plot:
```{r, eval = FALSE}
ggplot(wpdata.pred, aes(x=temp, y = depth)) + 
  geom_point() +
    geom_ribbon(aes(x=temp, ymin = lwr, ymax = upr, fill = "prediction"), alpha = .1) +
  geom_smooth(method = "lm", aes(fill = "confidence"), alpha = .4) +
  labs(x = "air temperature (C)", y= "nest depth (cm)", 
       title= "woodpeckers scatterplot", fill = "Type")
```


### Tools for displaying your model
As described in the Formatting Tables in Markdown (Section \@ref(markdown-tables)), you can use the package `stargazer` to create a nice table of model results in your pdf. The entire R chunk to do this in a `pdf` doc format is shown below (but not evaluated in this html book). You will need to add the R chunk option `results='asis'` to get the table formatted correctly. I also include the `message=FALSE` option in the chunk below that runs the library command to suppress the automatic message created when running the `library` command with `stargazer`. 

<pre><code>```{r, results='asis', message=FALSE}
library(stargazer)
stargazer(wood.lm, header=FALSE, single.row = TRUE, title = "SLR of depth on temp")
```</code></pre>

The `kable` function (from the `knitr` package) works well in all output environments (e.g. pdf, html, word). The input to this function needs to be a data frame, so we can use the `tidy` version of our model summary:

```{r, eval = FALSE}
library(knitr)
kable(tidy(wood.lm, conf.int = TRUE), digits = 3)
```



--------------------------------

## Checking model assumptions and fit {#slr-assump}



--------------------------------

## Example: SLR assumptions (day 4) {#slr-example2}