# Logistic Regression {#logistic}

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, prompt=TRUE, collapse = TRUE, message = FALSE, warning = FALSE)
```


This chapter covers material from chapters 20-21 of Sleuth. 

## The variables
Suppose we have a **categorical** response variable $Y$ that can take one of two values, which we will generically call a **success** or **failure**. We want to relate the **probability of success** to $p$ explantory variables (aka predictors, covariates) $x_1, \dotsc, x_p$. There is no restriction on the type of covariates, they can be both quantitative and categorical variables.

### Example: Donner party EDA {#logistic-donner1}
Sleuth case study 20.1 looks at data from the infamous [Donner party](https://en.wikipedia.org/wiki/Donner_Party). This wagon train was migrating to the west coast of the US in the mid-1800s and became snow-bound in the Sierra Nevada mountains during the winter of 1846-7. We are interested in  modeling the categorical variable "survival" (or not) as a function of individual covariates like age or sex. 

```{r}
library(Sleuth3)
donner <- case2001 
summary(donner)
```



A stacked bar graph shows that females had a higher surival rate than males:
```{r}
library(ggplot2)
ggplot(donner, aes(fill = Status, x = Sex)) + 
  geom_bar(position="fill") + 
  labs(y="proportion", title="Surival rates by Sex")
```

We can use the `dplyr` package's `group_by` function to divide the data into the two `Sex` groups and compute the proportion who `Survived` within each group. Here we see that  2/3 of females survived while only 1/3 of males survived. 
```{r}
library(dplyr)
donner %>% 
  group_by(Sex) %>%     # for each Sex group
  summarize(mean(Status == "Survived"))  # proportion who survived
```


A side-by-side boxplot shows that people who survived tended to be younger:
```{r}
library(ggplot2)
ggplot(donner, aes(x = Status, y = Age)) + 
  geom_boxplot() + 
  coord_flip()
```

We can get stats by status group:
```{r}
donner %>%
  group_by(Status) %>%   # for each status group
  summarize(mean(Age), sd(Age), median(Age))  # get summary stats
```

But these stats are not quite what we want when trying to model survival as a function of age. E.g. of people who survived, we know that the mean age was 27.2 while of people who died, the mean age was 35.5. But this is the wrong direction of conditioning, we would like to say how survival rates change as we increase age by a year, for example. We could employ a scatterplot for this purpose, but we need to recode (in `dplyr`) `Status` into a binary variable that recodes survival as a 1 and death as a 0:
```{r}
donner$Ind_surv <- recode(donner$Status, Survived = 1, Died = 0)
```
Then use a `jitter` plot to avoid overplotting:
```{r}
ggplot(donner, aes(x = Age, y = Ind_surv)) + 
  geom_jitter(aes(color=Status), height = .01) 
```

As we increase age, try vertically "slicing" these points. The proportion of survivals in these similar age groups will just be the mean of the binary 0's and 1's in each slice. We have more blue points than red in the low age slices than in high age, so we see that the probability of survival tends to decrease as age increases.  We will next consider how to construct a model that gives us a "best fit" curve for this probability of surival. 


## The Bernoulli distribution {#logistic-bernoulli}

The Bernoulli distribution is a probability model for a random trial that has two possible outcomes: success or failure. A Bernoulli **random variable** $Y$ "counts" the number of successes in a Bernoulli random trial. If a "success" occurred then $Y=1$ and if a "failure" occurred then $Y=0$. 

We will let $\pi$ be the probability of success:
$$
\pi = P(Y=1) = P(success), \ \ \ \ \ 1-\pi = P(Y=0) = P(failure) 
$$
If $Y$ is a Bernoulli random variable, then we can use the shorthand notation $Y \sim Bern(\pi)$ to denote this. 

The **expected value**, or mean, of $Y$ is equal to 
$$
E(Y) = \mu = \pi
$$
and the standard deviation of $Y$ is equal to 
$$
SD(Y) = \sigma = \sqrt{\pi(1-\pi)}
$$
The expected value (or mean) of a random variable measures the "long run" average value that we would see from $Y$ if we were to repeat the random trial many, many times. The standard deviation tells us how these values of $Y$ will vary over these repeated trials. 



## The logistic model form {#logistic-model}

The population, or data generating, model for a **logistic** regression model for $Y$ assumes that each $Y_i$ is a Bernoulli random variable whose probability of success **depends on covariates** $\pmb{x_{1,i}, \dotsc, x_{p,i}}$. Specifically,

- $Y_i \mid X_i \overset{indep.}{\sim} Bern(\pi(X_i))$
    - binary response: $Y_i$'s are categorical with only two options
    - independence: Given $X_i$ values, $Y_i$'s are independent

We connect the linear combination of predictors 
$$
\eta_i = \beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}
$$
to the probability of success using the logistic function form:
$$
\pi(X_i) = \dfrac{e^{\eta_i}}{1 + e^{\eta_i}} = \dfrac{e^{\beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}}}{1 + e^{\beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}}}
$$

This function form is used because its inverse is equal to 
$$
\eta_i =  \beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}=  \ln \left(  \dfrac{\pi(X_i)}{1-\pi(X_i)}\right)
$$
This function is called the **logit function**: $logit(\pi) = \ln \left(  \dfrac{\pi}{1-\pi}\right)$. This  means that a one unit increase in $x_1$ can be interpreted as an additive  $\beta_1$ change in in the logit function, holding other terms fixed.  But what does this mean?

The **odds of success** is defined as the ratio of the probability of success to the probability of failure:
$$
odds = \dfrac{\pi(X)}{1-\pi(X)}
$$
For example, if the probability of success is 0.6 then the odds of success is $0.6/0.4 = 1.5$. Meaning for every 6 successes, we see 4 failures. If the probability of success is 0.1, then the odds of success is $0.1/0.9 \approx 0.111$, meaning for every 1 success we see 9 failures. Odds greater than 1 indicate the probability of success is above 50% while odds less than 1 indicate the probability of success is less than 50%.

So, we now can see that the logit function equals the **log-odds of success**:
$$
\ln \left(  \dfrac{\pi(X_i)}{1-\pi(X_i)}\right) = \beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}
$$
This model form is an example of a **generalized linear model** which relates the response $Y$ to predictors through a linear combination $\eta$ of predictors. Is does this by defining the following functions:

- The **kernel mean function** defines the expected value (mean) of $Y$ as a function of $\eta$.
    - in a logistic model, the kernel mean function is the logistic function $E(Y \mid X) = \pi(X) = \dfrac{e^{\eta}}{1+e^{\eta}}$
- The **link function** defines the linear combination $\eta$ as a function of the mean of $Y$.
    - in a logistic model, the link function is the logit function $\eta = \ln(\pi/(1-\pi))$
- These two functions are inverses of one another. 


```{r, fig.height=4, fig.width=9, echo=FALSE}
par(mfrow=c(1,2))
curve(exp(x)/(1+exp(x)), from=-10, to=10, ylab="logistic(eta)", xlab="eta")
title("logistic")
curve(log(x/(1-x)), from=0, to=1,  xlab="p", ylab="logit(p)")
title("logit")
```

### Interpretation
Changes in predictors can be interpreted as changes in the **odds** of success (we can't make general statement about changes in the **probability** of success). Specifically, we  "unlog" the logit equation to get an expression for the odds of success for predictors $x_1, \dotsc, x_p$:
$$
odds(x_1, \dotsc, x_p) = \dfrac{\pi(X)}{1-\pi(X)} = e^{\beta_0 + \beta_1 x_{1} + \dotsm +  \beta_p x_{p}}
$$

What happens if we increase $x_1$ by one unit, holding other predictors fixed? 
$$
odds(x_1+1, \dotsc, x_p)  = e^{\beta_0 + \beta_1 (x_{1}+1) + \dotsm +  \beta_p x_{p}} = e^{\beta_0 + \beta_1 x_{1} + \dotsm +  \beta_p x_{p}} \times e^{\beta_1}
$$
Increasing $x_1$ by one unit has a **multiplicative** change of  $e^{\beta_1}$ in the **odds of success**. Note that this is a similar interpretation to the exponential model in SLR or MLR. 

The multiplicative change of $e^{\beta_1}$ is also called the **odds ratio** for a one unit increase in $x_1$. An odds *ratio* is just the ratio of the odds for two different groups, here groups with $x_1+1$ vs. $x_1$:
$$
\dfrac{\textrm{odds of succes at } x_1+1}{\textrm{odds of succes at } x_1} = \dfrac{odds(x_1+1, \dotsc, x_p) }{odds(x_1, \dotsc, x_p) } = e^{\beta_1}
$$

What is we have a predictor that is logged? 
$$
odds(x_1, \dotsc, x_p) = e^{\beta_0 + \beta_1 \ln(x_{1}) + \dotsm +  \beta_p x_{p}} =   e^{\beta_0}x_1^{\beta_1} e^{\beta_2 x_2 + \dotsm +  \beta_p x_{p}}
$$

Then our interpretation is similar to a power model. Changing $x_1$ by a factor of $m$:
$$
odds(mx_1, \dotsc, x_p) =    e^{\beta_0}(mx_1)^{\beta_1} e^{\beta_2 x_2 + \dotsm +  \beta_p x_{p}} =    e^{\beta_0}x_1^{\beta_1} e^{\beta_2 x_2 + \dotsm +  \beta_p x_{p}} \times m^{\beta_1}
$$
results in a multiplicative change of $m^{\beta_1}$ in the odds of success.

## Inference and estimation

Estimation of logistic model parameters $\beta_0, \dotsc, \beta_p$ is done using **maximum likelihood** estimation (MLE). The likelihood function is the probability of the observed data, writen as a function of our unknown $\beta$'s *(which are used to compute $\pi(X_i)$'s)
$$
L(\beta) = \prod_{i=1}^n \pi(X_i)^{y_i} (1-\pi(X_i))^{1-y_i}
$$
Notice that for each case $i$, the term in this product is equal to just $\pi(X_i)$ when case $i$ is a success ($y_i=1$) and equal to $1-\pi(X_i)$ when case $i$ is a failure ($y_i=0$).

Our MLE method says to find the $\beta$'s that maximize the likelihood $L(\beta)$ of the observed data. Unlike SLR or MLR, there is no "closed form" for these MLE $\hat{\beta}_i$ estimates (meaning we can't write down a formula for the estimates). Rather, software uses a numerical optimization method to compute the MLEs $\hat{\beta}_i$ **and** the standard errors  $SE(\hat{\beta}_i)$. (The R function `glm` uses Iterative reweighted least squares.)

These MLE estimates of $\beta$ parameters are approximately normally distributed and unbiased when n is "large enough." Much like with "intro stats" inference, when your response variable is categorical (or, equivalently, binary 0/1), we usually use the normal distribution for inference (CI and tests). When your response variable is quantitative (like in SLR/MLR models), we usually use the t-distribution for inference. 

### Confidence intervals for $\pmb{\beta_i}$
A $C$% confidence interval for $\beta_i$ equals
$$
\hat{\beta}_i \pm z^*SE(\hat{\beta}_i)
$$
where $z^*$ is the   $(100-C)/2$ percentile from the $N(0,1)$ distribution.

### Hypothesis tests for $\pmb{\beta_i}$
We can test the hypothesis
$$
H_0: \beta_i = \beta^*_i
$$
with the following z-test statistic:
$$
z =\dfrac{\hat{\beta}_i - \beta^*_i}{SE(\hat{\beta}_i)}
$$
where $\beta^*_i$ is our hypothesized value of $\beta_i$ . The $N(0,1)$ is used to compute the p-value that is appropriate for whatever $H_A$ is specified. 

The usual test results given by standard regression output tests whether a parameter value (intercept or slope) is equal to 0 vs. not equal to 0:
$$
H_0: \beta_i = 0 \ \ \ \ \ H_A: \beta_i \neq 0
$$
with a test stat of 
$$
z =\dfrac{\hat{\beta}_i - 0}{SE(\hat{\beta}_i)}
$$


### R `glm` 
We fit a logistic regression model in R with the `glm` function. The basic syntax is
```
glm(y ~ x1 + x2, family = binomial, data= )
```
Careful not to forget the `family=binomial` argument! If you omit this, you will just be trying to fit a regular MLR model which is not appropriate for a categorical response. 

The variable `y` can be either form:

- `y` can be binary 0/1 coded response where `1` is a "success"
- `y` can be a `factor` variable with two levels. The **second** level is what R will call a "success"

Once you fit a `glm` model, you can extract attributes of the model

- `fitted(my.glm)` gives the estimated probabilities of success for each case in your data
- `predict(my.glm)` gives estimated log-odds of success for each case in your data. Add `newdata=` to get predicted log-odds for new data.
- `predict(my.glm, type = "response")` gives estimated probabilities of success for each case in your data. Add `newdata=` to get predicted log-odds for new data.

The `broom` package also allows us to get fitted probabilities or log odds for all cases in the data, or for new data:

- `augment(my.glm)` gets estimated log-odds of success added to the variables used in the `glm` fit. 
    - add `data=my.data` to get estimated log-odds added to the full data set `my.data` used in the `glm` fit
    - add `newdata= new.data` to get predicted log-odds added to the new data set `new.data`
- `augment(my.glm, type.predict= "response")` gets estimated probabilities of success added to the variables used in the `glm` fit. 
    - add `data=my.data` to get estimated probabilities added to the full data set `my.data` used in the `glm` fit
    - add `newdata= new.data` to get predicted probabilities added to the new data set `new.data`

### Example: Donner party model {#logistic-donner2}
Let's revist the Donner party data and start with considering the logistic regression of survival status on age (only).  We can add the fitted logistic model probability curve the scatterplot we created in Section \@ref(logistic-donner1). We use the `glm` smoothing method with an `args` that specifies the `binomial` family:

```{r}
ggplot(donner, aes(x = Age, y = Ind_surv)) + 
  geom_jitter(aes(color=Status), height = .01) + 
  geom_smooth(method="glm", method.args = list(family=binomial), se=FALSE) + 
  labs(y = "probability of survival")
```

We can see that an age of about 30 yields an estimated survival probability of 50% while an age of about 45 yields an estimated survival probability of 25%. We can better quantify these values by fitting the model using the `glm` function.

Here is our simple model, the logistic regression of survival on age:
$$
logit(\pi) = \log(\dfrac{\pi}{1-\pi}) = \beta_0 + \beta_1 Age 
$$

```{r}
levels(donner$Status)  # second level = Surived
donner.glm1 <- glm( Status ~ Age , family=binomial, data=donner)
summary(donner.glm1)
confint(donner.glm1)
```

The estimated log odds of survival is 
$$
logit(\hat{\pi}) = \dfrac{\hat{\pi}}{1-\hat{\pi}} = 1.81852  -0.06647  Age
$$
and the estimated odds of survival is 
$$
\hat{odds}(age) = \dfrac{\hat{\pi}}{1-\hat{\pi}} = e^{1.81852}e^{-0.06647 Age}
$$
and the estimated probability of survival is 
$$
\hat{\pi}(age) =  \dfrac{e^{1.81852  -0.06647  Age}}{1+e^{1.81852  -0.06647  Age}}
$$

A one year increase in age  will have a $e^{-0.06647} = 0.936$ multiplicative change on the odds of survival. A one year increase in age decreases the odds of survival by 6.4% (95% CI 0.3% to 12.2%).  
```{r}
exp(-0.06647)   # factor change
100*(exp(-0.06647) - 1)  # percent change
exp(-0.06647 + c(-1,1)*qnorm(0.975)*0.03222)   # factor change CI
100*(exp(-0.06647 + c(-1,1)*qnorm(0.975)*0.03222) - 1)  # % change CI
```

The `broom` package's `tidy` function can also be used to get estimates, SEs and confidence intervals. If we add `exponentiate=TRUE`, then we we get exponentiated estiamtes and confidence intervals (but SE, test stat and p-values are untouched). 
```{r}
library(broom)
tidy(donner.glm1, conf.int=TRUE)
tidy(donner.glm1, conf.int=TRUE, exponentiate = TRUE)
```

We can use the `predict` command with `response` type values to get predicted survival rates for 30 and 45 year olds:
```{r}
new.ages <- data.frame(Age = c(30, 45))
predict(donner.glm1, newdata = new.ages, type="response")
exp(1.81852 - 0.06647*30)/(1+exp(1.81852 - 0.06647*30))  # prob age=30
exp(1.81852 - 0.06647*45)/(1+exp(1.81852 - 0.06647*45))  # prob age=45
```
So we have
$$
\hat{\pi}(age = 30)  =  \dfrac{e^{1.81852  -0.06647(30)}}{1+e^{1.81852  -0.06647(30)}} \approx 0.456
$$
and
$$
\hat{\pi}(age = 45)  =  \dfrac{e^{1.81852  -0.06647(45)}}{1+e^{1.81852  -0.06647(45)}} \approx 0.236
$$

Finally, what if we want to understand how the odds of *death* change as a function of age? Well, odds of death is equal to the ratio death to survival probabilities:
$$
\hat{odds.death}(age) = \dfrac{1-\hat{\pi}}{\hat{\pi}} = \dfrac{1}{e^{1.81852}e^{-0.06647 Age}} = e^{-1.81852}e^{0.06647 Age}
$$
A one year increase in age  will have a $e^{0.06647} = 1.069$ multiplicative change on the odds of survival. A one year increase in age increases the odds of death by 6.9% (95% CI 0.3% to 13.8%).  
```{r}
exp(0.06647)   # factor change in odds of death
100*(exp(0.06647) - 1)  # percent change
exp(0.06647 + c(-1,1)*qnorm(0.975)*0.03222)   # factor change CI
100*(exp(0.06647 + c(-1,1)*qnorm(0.975)*0.03222) - 1)  # % change CI
```

We can verify our mathematical work by refitting a `glm` with an indicator of *death*:
```{r}
donner$Ind_death <- recode(donner$Status, Survived = 0, Died = 1)
tidy(glm(Ind_death ~ Age, family = binomial, data=donner))
```

### Example: Donner party, adding sex {#logistic-donner3}
We will fit the  logistic regression of survival on age and sex:
$$
logit(\pi) = \log(\dfrac{\pi}{1-\pi}) = \beta_0 + \beta_1 Age + \beta_2 Male
$$

The estimated mode fit is:
```{r}
donner.glm2 <- glm( Status ~ Age + Sex, family=binomial, data=donner)
donner.glm2
tidy(donner.glm2, conf.int=TRUE)
tidy(donner.glm2, conf.int=TRUE, exponentiate = TRUE)
```

The estimated log odds of survival is 
$$
logit(\hat{\pi}) = \dfrac{\hat{\pi}}{1-\hat{\pi}} = 3.23041 -0.07820  Age -1.59729 Male
$$
and the estimated odds of survival is 
$$
odds(Sex, Age) = \dfrac{\hat{\pi}}{1-\hat{\pi}} = e^{3.23041}e^{-0.07820 Age}e^{-1.59729 Male} = (25.3)(0.925)^{Age}(0.202)^{Male}
$$
and the estimated probability of survival is 
$$
\hat{\pi}(Sex, Age) =  \dfrac{e^{3.23041 -0.07820  Age -1.59729 Male}}{1+e^{3.23041 -0.07820  Age -1.59729 Male}} 
$$

The exponentiated coefficient estimates give the odds ratios for a one unit increase in age or for males (compated to females). 

Holding gender constant, a one year increase in age decreases the odds of survival by 7.5% (95% CI 0.5% to 14.0%).  
```{r}
exp(-0.07820)  # age effect on odds
100*(exp(-0.07820) - 1)  # % change
exp(-0.07820 + c(-1,1)*qnorm(0.975)*0.03728)  # CI for factor 
100*(exp(-0.07820 + c(-1,1)*qnorm(0.975)*0.03728) - 1)  # CI for % change
```

Holding age constant, males had a 79.8% lower odds of survival compared to females (95% CI 11.0% to 95.4%). 
$$
\textrm{estimated odds ratio of survival for males vs females} = \dfrac{\hat{odds}(Sex=male,Age)}{\hat{odds}(Sex=female,Age)} = e^{-1.5973}
$$


```{r}
exp(-1.59729)
100*(exp(-1.59729) -1 )
exp(-1.59729 + c(-1,1)*qnorm(0.975)*0.75547)
100*(exp(-1.59729 + c(-1,1)*qnorm(0.975)*0.75547) - 1)
```

What if we wanted the odds ratio for comparing females to males? 
$$
\textrm{estimated odds ratio of survival for females vs males} = \dfrac{\hat{odds}(Sex=female,Age)}{\hat{odds}(Sex=male,Age)} = \dfrac{1}{e^{-1.5973}} = e^{1.5973}
$$
Holding age constant, females had a 4.9-fold increased odds of survival (95% CI 1.1 to 21.7).
```{r}
# odds ratio Female/Male
1/exp(-1.59729)  
1/exp(-1.59729 + c(-1,1)*qnorm(0.975)*0.75547)
```

There are two separate log-odds, odds and probability functions for the two levels of `Sex` in the model. Since we do not have an interaction between `Sex` and `Age`, we have a parallel line log-odds model. To plot the probabilities for each `Sex` as `Age` varies, we need to get the fitted probabities for each case in the data set:
```{r}
library(broom)
donner.aug2 <- augment(data=donner, donner.glm2, type.predict = "response")
head(donner.aug2)
```

Then add `geom_line` with the `.fitted` probabilities for each `Sex` (by `linetype`) added to the y-axis of the jittered scatterplot:
```{r}
ggplot(donner.aug2, aes(x = Age, y = Ind_surv)) + 
  geom_jitter(aes(color=Status), height = .01) + 
  geom_line(aes(y=.fitted, linetype=Sex)) + 
  labs(y = "probability of survival")
```

Note that these lines are *not* parallel on the probability scale while they are on the log-odds scale. This is because we apply the logistic function to the log-odds to produce our probabilities. The logistic function is not a linear function, so it does not produce parallel lines (or even lines at all). 