# Poisson Regression {#poisson}

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, prompt=TRUE, collapse = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(broom)
library(gridExtra)
```


This chapter covers material from chapter 23 of Sleuth. 

## The Poisson distribution {#pois-dist}

The Poisson distribution is a probability model for a random variable that counts the number of "events" (or successes) that occur in a fixed period of time and/or space. For example:

- the number of accidents that occur *per month* at an intersection 
- the number of cancer cases *per year* in a county
- the number of potholes *per mile* on interstate 35

To model these types of counts with a Poisson distribution, we must assume that

- two events can't occur at exactly the same time/location (e.g. two different accidents can't occur at exactly the same time)
- events occur independently over time/space (e.g. the occurance of one accident isn't going to make another more or less likely)

If $Y$ is a random variable with a Poisson distribution, then the probability that it equals some number $y$ is
$$
P(Y = y) = \dfrac{e^{\lambda}\lambda^y}{y!}  \ \ \textrm{ for any } y = 0,1,2,\dotsc
$$
Note that Poisson counts $Y$ don't have an "upper limit" like Binomial counts do so it can take on any integer value from 0 on up. 

The parameter $\lambda$ ("lambda") in this model tells us the *mean* response value, or the expected number of events per unit of time/space:
$$
E(Y) = \mu = \lambda
$$
We need to have $\lambda >0$ since our counts can't take on negative values. The variance of $Y$ is *also* equal to the mean:
$$
Var(Y) = \sigma^2 = \lambda
$$ 
and the standard deviation is then equal to 
$$
SD(Y) = \sigma = \sqrt{\lambda}
$$

So the expected rate of events $\mu = \lambda$  also tells us how variable our observed counts are likely to be. The larger the mean rate of events, the more variable our counts will be. Hence, our Poisson regression model will *not* have a "constant variance" assumption. The Poisson distribution is also skewed right for small values of the mean rate but becomes more symmetric as $\mu$ gets bigger. The plots below display Poisson distributions for a few values of $\mu$.


```{r, echo=FALSE}
set.seed(52115)
plotA <- ggplot(data.frame(x=0:6), aes(x,y=dpois(x,1))) + geom_segment(aes(xend=x,yend=0)) + labs(title="mu = 1")
plotB <- ggplot(data.frame(x=0:10), aes(x,y=dpois(x,2))) + geom_segment(aes(xend=x,yend=0))+ labs(title="mu = 2")
plotC <- ggplot(data.frame(x=0:20), aes(x,y=dpois(x,6))) + geom_segment(aes(xend=x,yend=0)) + labs(title="mu = 6")
plotD <- ggplot(data.frame(x=0:33), aes(x,y=dpois(x,16))) + geom_segment(aes(xend=x,yend=0)) + labs(title="mu = 16")

grid.arrange(plotA,plotB,plotC,plotD)
```


## The Poisson model form {#logistic-model}

We will assume that our response $Y_i$ for case $i$ is modeled by a Poisson distribution with mean (expected rate) that we will denote as $\mu(Y_i \mid X_i)$ (instead of $\lambda$, just to be consistent with earlier models):
$$
Y_i \mid X_i \overset{indep.}{\sim} Pois(\mu(Y_i \mid X_i))
$$

Our modeling goal is once again to model the mean response $\mu(Y_i \mid X_i)$ as a function of predictor $X_i = (x_{1,i}, \dotsc, x_{p,i})$. To do this we need to ensure that we have a model that always maps values of $X_i$ to *positive* values of $\mu(Y_i \mid X_i)$. We will do this by using another **generalized linear model**. 

The **link function** for our model is the log function defines the linear combination of predictors as the log of the mean response:
$$
\ln(\mu(Y_i \mid X_i)) = \eta_i = \beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}
$$
The **kernel mean function** for our model, which is the inverse of the link, is the exponential function so that 
$$
\mu(Y_i \mid X_i)  = e^{\eta_i} =e^{\beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}}
$$
This ensures that for any value of $\eta_i$, we get a positive value for our mean response. This is an example of a *log-linear* model.

The scatterplots below show an example of data generated from a SLR model and a Poisson model. Notice that the mean line in a SLR is a linear function of $x$ and we see symmetric, constant variation around the mean line. In the Poisson model data, we see a non-linear increase in the mean response as $x$ grows, along with increasing variation around the mean line. The variation around the mean line is right-skewed for smaller mean values but becomes more symmetric for larger mean values. 
```{r, echo=FALSE}
set.seed(1290482)
x <- runif(1000, 0, 20)
y <- rpois(1000, exp(-1.7 + .2*x))
plotA <- ggplot(data.frame(x,y), aes(x=x, y=y)) + 
  geom_point() + 
  stat_function(fun=function(x) {exp(-1.7+.2*x)}, color="blue", size=2) + 
  labs(title="Poisson model")
y <- rnorm(1000, -1.7 + 2*x, sd=10)
plotB <- ggplot(data.frame(x,y), aes(x=x, y=y)) + 
  geom_point() + 
  stat_function(fun=function(x) {-1.7 + 2*x}, color="blue", size=2) + 
  labs(title="SLR model")
grid.arrange(plotB, plotA, nrow=2)
```

### Interpretation

If we have untransformed predictors, we can say that a 1 unit increase in $x_1$, for example, is associated with a $e^{\beta_1}$ *multiplicative* change in the **mean response** $\mu$:
$$
\mu(Y \mid X+1)  = e^{\beta_0 + \beta_1 (x_{1}+1) + \dotsm +  \beta_p x_{p}} = \mu(Y \mid X)e^{\beta_1} 
$$

Notice that this is similar to interpretation of an exponential model in SLR/MLR (Section \@ref(slr-trans)), except that in the Poisson model we are setting our linear combination of terms equal to the **logged-mean response** $\ln(\mu_Y \mid X)$. In the SLR/MLR model, we are setting our linear combination of terms equal to the **mean of the logged-responses** $\mu(\ln(y) \mid X)$. Interpretation of the Poisson log-linear model is easier because exponentiating the linear combination returns to us the the *mean* response. 

If we took the natural log of $x_1$, we can say that a multiplicative change of $m$  in $x_1$ is associated with a $m^{\beta_1}$ *multiplicative* change in the **mean response** $\mu$:
$$
\mu(Y \mid mX)  = e^{\beta_0 + \beta_1 \ln(mx_{1}) + \dotsm +  \beta_p x_{p}} = \mu(Y \mid X)m^{\beta_1} 
$$

## EDA

The Poisson model assumes that the quantitative predictors are linearly related to the log of the mean response. EDA to check this assume consists of a scatterplot of $\ln(y)$ vs. $x$, looking for a transformation of $x$, if needed, that displays a linear relationship. Constant variance is not assumed. 

### Example: Possums

The data in `possums` was collected from a sample of  $n = 151$ 3-hectare sites in Australia. Our goal is to determine which factors are associated with good habitat for possums. The response recored for each site is `y`= the number of possum species found on the site. We will start by just considering the predictors  `Bark` = bark quality index (low to high quality).
```{r}
possums <- read.csv("http://people.carleton.edu/~kstclair/data/possums.csv")
summary(possums$y)
summary(possums$Bark)
```
At least 25% of sites had no possum species observed, and about 75% of sites had 2 or fewer species observed. The middle 50% of sites had bark indices between 6 and 10. 

To check if `Bark` needs to be transformed, we need look at log-`y` vs `Bark`. Since `y` has many 0 species counts we will need to add a small amount (.5) to the vector `y` before plotting.  As bark index increases, there is a slightly non-linear increase in number of species observed for untransformed `Bark`. The plot of `log(y)` vs. `log(bark)` looks  more linear. A small amount of jitter was added in both directions to avoid overplotting issues.
```{r}
plotA <- ggplot(possums, aes(x=Bark, y = y + .5)) + 
  geom_jitter(width = .01, height=.01) +  geom_smooth(se=FALSE) + 
  scale_y_log10() + 
  labs(title="Log-Number species vs. Bark index")

plotB <- ggplot(possums, aes(x=Bark, y = y + .5)) + 
  geom_jitter(width = .01, height=.01) + geom_smooth(se=FALSE) + 
  scale_y_log10() + scale_x_log10() + 
  labs(title="Log-number species vs. log-Bark index")


grid.arrange(plotA, plotB, nrow=1)
```


## Inference and estimation {#poisson-est}

Estimation of Poisson model parameters $\beta_0, \dotsc, \beta_p$ is done using **maximum likelihood** estimation (again!). The likelihood function is the probability of the observed data, writen as a function of our unknown $\beta$'s where here $\mu(X_i)  = e^{\beta_0 + \beta_1 x_{1,i} + \dotsm +  \beta_p x_{p,i}}$ 
$$
L(\beta) = \prod_{i=1}^n \dfrac{e^{\mu(X_i)}\mu(X_i)^{y_i}}{y_i!} 
$$


Like other GLMs, these MLE estimates of $\beta$ parameters are approximately normally distributed and unbiased when n is "large enough" or when $\mu(X_i)$'s are "large enough." 

### Confidence intervals for $\pmb{\beta_i}$
A $C$% confidence interval for $\beta_i$ equals
$$
\hat{\beta}_i \pm z^*SE(\hat{\beta}_i)
$$
where $z^*$ is the   $(100-C)/2$ percentile from the $N(0,1)$ distribution. To get the multiplicative change in the mean response, we just exponentiate the CI:
$$
e^{\hat{\beta}_i \pm z^*SE(\hat{\beta}_i)}
$$

### Hypothesis tests for $\pmb{\beta_i}$
We can test the hypothesis
$$
H_0: \beta_i = \beta^*_i
$$
with the following z-test statistic:
$$
z =\dfrac{\hat{\beta}_i - \beta^*_i}{SE(\hat{\beta}_i)}
$$
where $\beta^*_i$ is our hypothesized value of $\beta_i$ . The $N(0,1)$ is used to compute the p-value that is appropriate for whatever $H_A$ is specified. 

The usual test results given by standard regression output tests whether a parameter value (intercept or slope) is equal to 0 vs. not equal to 0:
$$
H_0: \beta_i = 0 \ \ \ \ \ H_A: \beta_i \neq 0
$$
with a test stat of 
$$
z =\dfrac{\hat{\beta}_i - 0}{SE(\hat{\beta}_i)}
$$


### R `glm` 
We fit a Poisson regression model in R with the `glm` function. The basic syntax is
```
glm(y ~ x1 + x2, family = poisson, data= )
```
Careful not to forget the `family=poisson` argument! If you omit this, you will just be trying to fit a regular MLR model which is not appropriate for a categorical response. 

Once you fit a `glm` model, you can extract attributes of the model

- `fitted(my.glm)` gives the estimated mean response $\hat{\mu}_i$ for each case in your data
- `predict(my.glm)` gives estimated **log-mean** $\hat{\eta}_i$ for each case in your data. Add `newdata=` to get predicted log-mean for new data.
- `predict(my.glm, type = "response")` gives estimated **mean** $\hat{\mu}_i$  for each case in your data. Add `newdata=` to get predicted means for new data.

The `broom` package also allows us to get fitted probabilities or log odds for all cases in the data, or for new data:

- `augment(my.glm)` gets estimated **log-mean** $\hat{\eta}_i$ added to the variables used in the `glm` fit. 
    - add `data=my.data` to get estimated log-means added to the full data set `my.data` used in the `glm` fit
    - add `newdata= new.data` to get predicted log-means added to the new data set `new.data`
- `augment(my.glm, type.predict= "response")` gets estimated **mean** $\hat{\mu}_i$  added to the variables used in the `glm` fit. 
    - add `data=my.data` to get estimated means added to the full data set `my.data` used in the `glm` fit
    - add `newdata= new.data` to get predicted means added to the new data set `new.data`


### Example: Possums {#pois-ex2}

Our EDA above used `log(y)` just to explore whether we needed to transform our predictors bark. (Since log of our response should be linearly related to our predictor in Poisson regression.) But our Poisson model is just fit with `y`, not `log(y)` since our Poisson model already assumes log-linearity. We will fit the following Poisson regression of number of species on log-bark index:
$$
\ln(\mu(Y \mid x)) = \beta_0 + \beta_1 \ln(Bark)
$$

The model fit is below. 
```{r}
pos.glm <- glm(y ~ log(Bark), family=poisson, data=possums)
summary(pos.glm)
```

The estimated log mean function is
$$
\ln(\hat{\mu}(y \mid x)) =  -0.8801 + 0.5945 \ln(Bark)
$$
and the estimated mean function is 
$$
\hat{\mu}(Y \mid x) = e^{-0.8801 + 0.5945 \ln(Bark)} =  e^{-0.8801}Bark^{0.5945}
$$
The estimated mean number of possums per 3-hectare when Bark quality is 10 is 1.63 species:
$$
\hat{\mu}(Y \mid x=10) =   e^{-0.8801}10^{0.5945} =1.6302
$$
```{r}
predict(pos.glm, newdata=list(Bark=10))  # log mean
exp(predict(pos.glm, newdata=list(Bark=10)))  # mean 
predict(pos.glm, newdata=list(Bark=10), type="response")  # mean
```

With $n=151$, we have a large enough sample size to trust normal-inference methods for our MLE estimates. If $n$ was smaller, you would want to check summaries of the estimated mean values. Here we see that they range from an estimated 0.4 species per plot to 3.1 species per plot. 
```{r}
summary(fitted(pos.glm))
```

Bark index has a statistically significant effect on the mean number of species (z=4.45, p < 0.0001). Doubling bark index is associated with a 51% increase in the estimated mean number of species per 3-hectare plot (95% CI 26% to 81%).
$$
2^{0.5945}  = `r 2^(0.5945)` \ \ \ 2^{0.5945 + c(-1,1)*qnorm(.975)*0.1335} = `r 2^(0.5945 + c(-1,1)*qnorm(.975)*0.1335)`
$$

```{r}
2^(.5945)  # double bark index
100*(2^(.5945) - 1)
# 95% CI for beta1
0.5945 + c(-1,1)*qnorm(.975)*0.1335
# 95% CI for factor change in mu (doubling bark)
2^(0.5945 + c(-1,1)*qnorm(.975)*0.1335)
100*(2^(0.5945 + c(-1,1)*qnorm(.975)*0.1335) - 1)
```

We can visualize the mean response $\hat{\mu}(Y \mid x) = e^{-0.8801 + 0.5945 \ln(Bark)}$ on a plot log-bark and number of species by first plotting (un-logged) `y` against the log of `Bark`. Then we add the `geom_smooth` with `glm` smoother method and a `poisson` family listed as the argument. Here we also include a smaller amount of `jitter` to avoid overplotting.
$as a function of Bark index
```{r}
ggplot(possums, aes(x=Bark, y=y)) +
  geom_jitter(width = .01, height=.1) + scale_x_log10() + 
  geom_smooth(method="glm", method.args = list(family=poisson), se=FALSE) + 
  labs(title="Poisson regression of # species on log-bark index", 
       y = "# of species", x="bark index")
```


## Deviance for Binomial responses {#pois-dev}


In Poisson  regression, the  deviance is equal to
$$
G^2 = 2[\ln L(\bar{\mu}) - \ln L(\hat{\mu}(X))] = 2\sum_{i=1}^n \left[ y_i \ln \left( \dfrac{y_i}{\hat{\mu}(X_i)} \right) - (y_i- \hat{\mu}(X_i))  \right]
$$
Note that cases with large $y_i$ can contribute the most to this summation of individual case-level deviance. 

Again, the two likelihoods that are used to construct residual deviance are:

- $L(\hat{\mu}(X)):$ likelihood of the data that plugs in estimates $\hat{\mu}(X_i)$ from the Poisson model.
- $L(\bar{\mu}):$ likelihood of the data that plugs in estimates $\bar{\mu} = y_i$, basing a case's "predicted" value soley on the response observed for that case. This again is called a **saturated** model and it will always have a higher likelihood than the logistic model: $L(\bar{\mu}) \geq L(\hat{\mu}(X))$. 

Deviance for binomial models can be used for two types of hypothesis tests:

### Drop-in-deviance test {#pois-did}

This test is used to compare two models, just like in logistic models. Details are in Section \@ref(logistic-dev). One difference between the logistic model and Poisson is the "large sample size" condition needed to trust the drop-in-deviance p-value approximation. In Poisson models, we need either a large sample size $n$ or $\hat{\mu}$'s that are large. 

Two compare two Poisson models using deviance, run
```
anova(pois.red, pois.full, test = "Chisq")
```
## Checking Assumptions {#pois-assump}

**Assumptions** in a Poisson model include 

- **Cases are independent:** Independence of cases takes an understanding of how the data was collected. 
- **Log-mean linearity:** Log-mean linearity can be checked with an a plot of log-response against quantitative predictors and residual plots. 

A **third assumption** is that given predictor values, the **counts of events $Y_i$ has a Poisson distribution**. This means that for each case $i$ we assume the following:

- **The events occur independently**, and there is no clustering of events across time or space. Clustering of event occurrences induces more variation in our responses than our Poisson model assumes. 

For example, if one rush hour accident makes more accidents more likely in that time period, then we could see big swings in monthly accident counts. Some months with no rush hour accidents will see low counts while months with a rush hour accident could see a big uptick in total accident counts. This idea is demonstrated in the code below with the `y.clus` variable (10 accidents per month in winter months and 0 in other months) which has the same mean as a Poisson vector of accident counts, but which has a much larger measure of SD.  

```{r}
# monthly accident counts from a Poisson with mean around 3:
y.pois <- c(5, 4, 1, 5, 3, 2, 6, 1, 4, 5, 0, 4)
mean(y.pois)
sd(y.pois)
# monthly accident counts clustered in winter with mean around 3:
y.clus <- c(10,10,10,0,0,0,0,0,0,0,0,10)
mean(y.clus)   # same average
sd(y.clus)     # much bigger variance than poisson counts!
```


- **We have enough explanatory power in our predictors to adequately model the mean response.** If we lack this, then we can't adequately explain variations in our observed counts and we our residual deviance will be very large. 

If one, or both, of these assumptions is violated, then it often induces **extra-Poisson variation** (a.k.a. **overdispersion**). This means that the actual variation in our response $SD(Y\mid X_i)$ is **larger** than the  Poisson SD of $\sqrt{\mu(X_i)}$ that our model assumes. So the response variation is more dispersed than what our model assumes, making our reported **standard errors and p-values too small** and we could be over-reporting statistical significance. 

We can use another goodness-of-fit test, when $\mu_i$ are large enough, to check our Poisson distribution assumption. If we do find evidence of lack-of-fit in our  model, then you should

- Check deviance residuals as case influence stats to see if an outlier(s) is  affecting GOF results. 
- Check the log mean form and see if transformations of quantitative predictors are needed
- If outliers and transformations aren't a concern, then consider an alternative model:
    - quasi-Poisson logistic model 
    - a model that allows for correlated trials (like a mixed-effects Poisson model) 
    
### Goodness-of-fit test {#pois-gof}

In our Poisson model, we are assuming that an observed count for case $i$ behaves like a Poisson random variable. We compare our Poisson model's estimated mean counts for each case to the observed counts from the saturated model described above. The saturated model will have the best "fit" (highest likelihood) of the two models, but if the "fit" (likelihood) of the logistic model is "close" then we can claim that the logistic model is adequate. This is the motivation behind a "goodness-of-fit" test. 

Our hypothese for the GOF test are:
$$
H_0: \textrm{Poisson model}
$$
vs. 
$$
H_A: \textrm{saturated model}
$$

The test statistic for this is equal to the **residual deviance of the Poisson model** (difference in the $H_A$ and $H_0$ likelihoods):
$$
G^2 = 2[\ln L(\bar{\mu}) - \ln L(\hat{\mu}(X))]
$$

If $H_0$ is true and the data does fit the Poisson model, then  when estimated $\hat{\mu}_i$'s are large, $G^2$ will have an approximate  **chi-square distribution** with $n - (p+1)$ (model) degrees of freedom. (Note that $n-(p+1)$ is the difference in the number of parameters in the two models.) The p-value is the probability of observing deviance larger than our model's value:
$$
p-value = 1-P(\chi^2 > G^2) = 1-pchisq(G^2, df=n-(p+1))
$$
The suggested rule of thumb for "large $\hat{\mu}_i$"" is that we want most $\hat{\mu}_i$'s to be at least 5. 

For a GOF test, you should consider the following interpretations for "large" or "small" p-values:

- **Do not reject the null:** (large p-value)
    - Your Poisson model is adequate.
    - You don't have a large enough sample size $n$ to have the *power* to detect inadequacies in your model.

- **Reject the null:** (small p-value)
    - You have outlier(s) that are inflating the residual deviance.
    - Your logistic model is inadequate.
        - Your mean model is inadequate, it is ill-fitting and transformations are needed
        - Extra-Poisson variation: your response counts aren't well modeled by a Poisson model. For each case:
            - events occurances are not independent, usually clustered in some way
            - your choice of predictors isn't sufficient (i.e. you are missing key explanatory variables)

### GOF alternative

If the GOF p-value approximation is suspect due to small estimated means, then one ad hoc method of assessing the "Poisson-ness" of the data would be to compare the sample mean and variances of cases with similar predictor values. Since a Poisson variable should have mean and variances that are similar in size, this can be used to detect if our counts are overdispersed. The basic algorithm is similar to that used in the BWCA empirical log-odds plot outlined in Section \@ref(#example-bwca1). 

1. Group cases into groups with similar predictor values
2. Within each group, compute the sample mean and sample variance of the observed counts `y`
3. Plot means against variances, if they following the y=x line then the Poisson assumption looks adequate.  


### Example: Possums

In Section \@ref(pois-ex2) we tested the statistical significance of log-Bark with a Wald z-test. We could have instead used a drop in deviance test for this one term. Hypotheses are the same but the mechanics of each are slightly different. Either method is ok since n=151 is fairly large so their conclusions should be similar. 

Using `anova` for this one-term model will test the significance of the one term, log-Bark, against the "null" model with only the intercept term. Here our test statistic is $G^2 = 19.979$ since this is the amount that the residual deviance decreases when adding log-Bark. The p-value, comptuted from a chi-square distribution with 1 degree of freedom, is extremely small indicating the strong significance of this term.
```{r}
anova(pos.glm, test="Chisq")
```

#### Goodness-of-fit

We might be tempted to take the deviance from our log-Bark model and run a GOF test to assess Poisson model adequacy:
```{r}
summary(pos.glm)
1-pchisq(167.51, df=149)  # p-value??
```

While the goodness of fit test gives a p-value of 0.1425, suggesting adequacy,  we should be suspect because most estimated means are small and well below our "large $\hat{\mu}_i$" threshold of 5.

```{r}
summary(fitted(pos.glm))
```

Our alternative method of checking whether our response is behaving like Poisson counts is the EDA check of sample means vs variances. We need to group cases by similar bark amounts. We can first note that the bark index is a rather discrete measure with many values (especially in mid-range values)
```{r}
table(possums$Bark)
```

We could simply try grouping by each individual bark value and getting the mean and variance of the observed counts for each. This will result in some 1 case bark values having a variance of `NA` (since we can't measure variability of one data value):
```{r}
pos.byBark <- possums %>% 
  group_by(Bark) %>%
  summarize(nY = n(), meanY = mean(y), varY = var(y))
pos.byBark %>% print(n=Inf)
```
Then plot mean count vs the variance of count. The means and variances for each value of bark are similar, so we have no graphical evidence of lack of fit.
```{r}
ggplot(pos.byBark, aes(x=meanY, y=varY)) + 
  geom_point() + 
  geom_abline(intercept=0, slope=1, linetype=2)
```

Alternatively, we could first group the cases according to the `ntile` function. Here we divide cases up into 12 similar groups based on bark, then get stats for each group and plot. A similar conclusion is made with this visual. No strong evidence of a trend in variance always being bigger (or smaller) than the mean count. 
```{r}
library(tidyverse)
pos.byBark2 <- possums %>% 
  mutate(Bark.grp = ntile(Bark, n=12)) %>%
  group_by(Bark.grp) %>%
  summarize(nY = n(), meanY = mean(y), varY = var(y))
pos.byBark2 %>% print(n=Inf)
ggplot(pos.byBark2, aes(x=meanY, y=varY)) + 
  geom_point() + 
  geom_abline(intercept=0, slope=1, linetype=2)
```

## Residuals and case influence for binomial responses {#pois-resid}

We have a few options for residuals for a binomial logistic model. The two most common residuals to consider are Pearson residuals and deviance residuals. 

**Pearson residuals** are basically response residuals standardized based on the binomial SD:
$$
pr_i = \dfrac{ y_i - \hat{\mu}(X_i)}{\sqrt{\hat{\mu}(X_i)}}
$$
We can get these residuals by requesting the "pearson" type of residual:

- `resid(my.glm, type = "pearson")`
- `augment(my.glm,  type.residuals = "pearson")` 


**Deviance residuals** are each case's contribution to the residual deviance, with a $\pm$ based on whether we over- or under-estimate a case's response (the $\pm$ is denoted by $sign(y_i - m_i\hat{\pi}(X_i))$):
$$
Dres_i =  sign(y_i - \hat{\mu}(X_i)) \sqrt{2 \left[ y_i \ln \left( \dfrac{y_i}{\hat{\mu}(X_i)} \right) - (y_i- \hat{\mu}(X_i))  \right] }
$$
We can get these residuals by requesting the default residual values:

- `resid(my.glm)`
- `augment(my.glm)` 

As with model GLMs: Pearson residuals are "easy" to interpret as the number of estimated SD's a response is from it's estimated mean. Deviance residuals are good to check if you find significant results in a GOF test. When $\hat{\mu}(X_i)$'s are large (at least 5), both types of residuals should be similar in value and have a $N(0,1)$ distribution (approximately). This means that most cases (~95%) should have residual values no more extreme than $\pm 2$.  Regardless of size of $m_i$, we should plot residuals vs. quantitative predictors to assess linearity of the log odds.

Case influence stats of leverage and Cook's distance can also be used to look for outliers. In a GLM, leverage measures both a cases's "extremeness" in terms of it's predictor values **and** it's extremeness in terms of it's **weight**. In a **Poisson** model, a case's weight is $\hat{\mu}(X_i)$. A case with higher values of $\hat{\mu}(X_i)$  are given more weight, and hence higher leverage, in the fitted model (e.g. in the estiamtes of $\hat{\beta}$). The value of Cook's distance also takes into account a cases leverage  and a case's residual value. 

You can get Pearson residuals, leverage and Cook's distance with `plot(my.glm, which = 5)`. You can also get leverage and Cook's distance vs. predictor values with `ggnostic(my.glm, columnsY = c(".hat",".cooksd"))` from the `GGally` package.